---
title: Exercies 1
output: pdf_document
toc: true
---

```{r setup, include=F}
require(tidyverse)
source('../helpers/graph_functions.R')
```

# Decimal Representations

#### 1.

Express $1/9$ as a repeating decimal, using a bar to indicate the repeating
digits. What are the decimal representations of $2/9$? $3/9$? $8/9$?

The pattern of numbers divded by 9.

$$
\begin{aligned}
\frac{1}{9} &= 0.\overline{111}\\\\
\frac{2}{9} &= 0.\overline{222}\\\\
\frac{3}{9} &= 0.\overline{333}\\
&\vdots\\
\frac{8}{9} &= 0.\overline{888}\\\\
\frac{9}{9} &= 1\\\\
\frac{10}{9} &= 1.\overline{111}\\\\
\frac{11}{9} &= 1.\overline{222}\\
&\vdots\\
\frac{18}{9} &= 2\\\\
\frac{19}{9} &= 2.\overline{111}\\
&\vdots
\end{aligned}
$$

#### 2.

Express $1/11$ as a repeating decimal, using a bar to indicate the repeating
digits. What are the decimal representations of $2/11$? $3/11$? $9/11$?
 
$$
\begin{aligned}
\frac{1}{11} &= 0.\overline{0909}\\\\
\frac{2}{11} &= 0.\overline{1818}\\\\
\frac{3}{11} &= 0.\overline{2727}\\\\
\frac{4}{11} &= 0.\overline{3636}\\
\vdots\\
\frac{9}{11} &= 0.\overline{8181}\\\\
\end{aligned}
$$

# Inequalities

#### 3.

If $2 < x < 6$, which of the following statements about $x$ are necessarily
true, and which are not necessarily true?

##### a.

$0 < x < 4$

> This is not necessarily true. The interval's minimum extends beyond the above
> interval's minimum.

##### b.

$0 < x - 2 < 4$

$$
2 < x < 6
$$

> This is necessarily true. The interval is equal to the above interval.

##### c.

$1 < \frac{x}{2} < 3$

$$
\begin{aligned}
1 < \frac{6}{x} < 3\\\\
\frac{1}{6} < \frac{1}{x} < \frac{1}{2}\\\\
6 > x > 2\\
2 < x < 6
\end{aligned}
$$

> This is necessarily true. When taking recipricals, we reverse the
> inequalities.

##### d.

$1 < \frac{6}{x} < 3$

$$
\begin{aligned}
\frac{1}{6} < \frac{1}{x} < \frac{1}{2}\\\\
6 > x > 2\\
2 < x < 6
\end{aligned}
$$

> This is necessarily true. When taking recipricals, we reverse the
> inequalities.

##### e.

$1 < \frac{6}{x} < 3$

$$
\begin{aligned}
1 < \frac{6}{x} < 3\\\\
\frac{1}{6} < \frac{1}{x} < \frac{1}{2}\\\\
6 > x > 2\\
2 < x < 6
\end{aligned}
$$

> This is necessarily true.

##### f.

$|x - 4| < 2$

$$
\begin{aligned}
-2 < x - 4 < 2\\\\
2 < x < 6
\end{aligned}
$$

> This is necessarily true.


##### g.

$-6 < -x < 2$

$$
\begin{aligned}
-6 < -x < 2\\
6 > x > -2\\
-2 < x < 6
\end{aligned}
$$

> This is not necessarily true. The interval's minimum extends past the above
> interval's minimum.

##### h.

$-6 < -x < -2$

$$
\begin{aligned}
-6 < -x < -2\\
6 > x > 2\\
2 < x < 6
\end{aligned}
$$

> This is necessarily true.

#### 4.

If $-1 < y - 5 < 1$, which of the following statements about $y$ are necessarily
true, and which are not necessarily true?

$$
\begin{aligned}
-1 < y - 5 < 1\\
4 < y < 6
\end{aligned}
$$

##### a.

$4 < y < 6$

> This is necessarily true.

##### b.

$-6 < y < -4$

> Not necessarily true.

##### c.

$y > 4$

> necessarily true. It follows from 4 < y < 6.

##### d.

$y < 6$

> necessarily true.

##### e.

$$
\begin{aligned}
0 < y - 4 < 2\\
4 < y < 6
\end{aligned}
$$

> necessarily true.

##### f.

$$
\begin{aligned}
2 < \frac{y}{2} < 3\\
4 < y < 6
\end{aligned}
$$

> necessarily true.

##### g.

$$
\begin{aligned}
\frac{1}{6} < \frac{1}{y} < \frac{1}{4}\\\\
6 > y > 4\\
4 < y < 6
\end{aligned}
$$

> necessarily true

##### h.

$$
|y - 5| < 1\\
-1 < y - 6 < 1\\
4 < y < 6
$$

> necessarily true

In Exercises 5--12, solve the inequalities and graph the solution sets.

#### 5.

$-2x > 4$

$$
\begin{aligned}
x < -8
\end{aligned}
$$

```{r problem 5, echo=F, fig.height=3, fig.width=3}
p5 <- Intervals(matrix(c(-20, -8), ncol=2), closed = c(F, F))
plot(p5, xlim=c(-10,-8))
```

#### 6.

$8 - 3x \ge 5$

$$
\begin{aligned}
8 &\ge 5 + 3x\\
3 &\ge 3x\\
1 &\ge x
\end{aligned}
$$

```{r problem 6, echo=F, fig.height=3, fig.width=3}
p6 <- Intervals(matrix(c(1, 10), ncol=2), closed = c(T, F))
plot(p6, xlim=c(1, 4))
```

#### 7.

$5x - 3 \le 7 - 3x$

$$
\begin{aligned}
5x - 3 &\le 7 - 3x\\
8x &\le 10\\
x &\le \frac{5}{4}
\end{aligned}
$$

```{r problem 7, echo=F, fig.height=3, fig.width=3}
p7 <- Intervals(matrix(c(-10, 5/4), ncol=2), closed = c(F, T))
plot(p7, xlim=c(-1, 5/4))
```

#### 8.

$3(2 - x) > 2(3 + x)$

$$
\begin{aligned}
3(2 - x) &> 2(3 + x)\\
6 - 3x &> 6 + 2x\\
0 &> 5x\\
x &< 0
\end{aligned}
$$

```{r problem 8, echo=F, fig.height=3, fig.width=3}
p8 <- Intervals(matrix(c(-10, 0), ncol=2), closed = c(F, F))
plot(p8, xlim=c(-1, 0))
```

#### 9.

$2x - \frac{1}{2} \ge 7x + \frac{7}{6}$

$$
\begin{aligned}
2x - \frac{1}{2} &\ge 7x + \frac{7}{6}\\
12x - 3 &\ge 42x + 7\\
-30x &\ge 10\\
x &\le -\frac{1}{3}
\end{aligned}
$$

```{r problem 9, echo=F, fig.height=3, fig.width=3}
p9 <- Intervals(matrix(c(-10, -1/3), ncol=2), closed = c(F, T))
plot(p9, xlim = c(-1, -1/3))
```

#### 10.

$\frac{6 - x}{4} < \frac{3x - 4}{2}$

$$
\begin{aligned}
6 - x &< 6x - 8\\
-7x &< -14\\
x &> 2
\end{aligned}
$$

```{r problem 10, echo=F, fig.height=3, fig.width=3}
p10 <- Intervals(matrix(c(2, 10), ncol=2), closed = c(F, F))
plot(p10, xlim = c(2, 5))
```

#### 11.

$\frac{4}{5}(x - 2) < \frac{1}{3}(x - 6)$

$$
\begin{aligned}
12(x - 2) &< 5(x - 6)\\
12x - 24 &< 5x - 30\\
7x &< -6\\
x &< -\frac{6}{7}
\end{aligned}
$$

```{r problem 11, echo=F, fig.height=3, fig.width=3}
p11 <- Intervals(matrix(c(-10, -6/7), ncol=2), closed = c(F, F))
plot(p11, xlim = c(-5, -6/7))
```

#### 12.

$-\frac{x + 5}{2} \le \frac{12 + 3x}{4}$

$$
\begin{aligned}
-2x - 10 &\le 12 + 3x\\
-5x &\le 22\\
x &\ge -\frac{22}{5}
\end{aligned}
$$

```{r problem 12, echo=F, fig.height=3, fig.width=3}
p12 <- Intervals(matrix(c(-22/5, 10), ncol=2), closed = c(T, F))
plot(p12, xlim=c(-22/5, 1))
```

# Absolute Value

Solve the equations in Exercises 13--18.

#### 13.

$|y| = 3$

$$
y = \pm3
$$

#### 14.

$|y - 3| = 7$

$$
\begin{aligned}
&y - 3 = 7 &\ \ & y - 3 = -7&\\
&y = 10 &\ \ & y = -4&
\end{aligned}
$$

#### 15.

$|2t + 5| = 4$

$$
\begin{aligned}
&2t + 5 = 4 &\ \ & 2t + 5 = -4&\\
&t = -\frac{1}{2} &\ \ & t = -\frac{9}{2}&
\end{aligned}
$$

#### 16.

$|1 - t| = 1$

$$
\begin{aligned}
&1 - t = 1 &\ \ & 1 - t = -1&\\
&t = 0 &\ \ & t = 2&
\end{aligned}
$$

#### 17.

$|8 - 3x| = \frac{9}{2}$

$$
\begin{aligned}
&8 - 3s = \frac{9}{2} &\ \ & 8 - 3s = -\frac{9}{2}&\\
&16 - 6s = 9 &\ \ & 16 - 6s = -9&\\
&s = \frac{7}{6} &\ \ & s = \frac{25}{6}&
\end{aligned}
$$

#### 18.

$\left|\frac{s}{2} - 1\right| = 1$

$$
\begin{aligned}
&\frac{s}{2} - 1 = 1 &\ \ & \frac{s}{2} - 1 = -1&\\
&s = 4 &\ \ & s = 0&
\end{aligned}
$$

Solve the inequalities in Exercises 19--34, expressing the solution sets as
intervals or unions of intervals. Also, graph each solution set on the real
line.

#### 19.

$|x| < 2$

$$
\begin{aligned}
-2 < x < 2\\\\
(-2, 2)
\end{aligned}
$$

```{r problem 19, echo=F, fig.height=3, fig.width=3}
p19 <- Intervals(matrix(c(-2, 2), ncol=2), closed = c(F, F))
plot(p19)
```

#### 20.

$|x| \le 2$

$$
\begin{aligned}
-2 \le x \le 2\\\\
[-2, 2]
\end{aligned}
$$

```{r problem 20, echo=F, fig.height=3, fig.width=3}
p20 <- Intervals(matrix(c(-2, 2), ncol=2), closed = c(T, T))
plot(p20)
```

#### 21.

$|t - 1| \le 3$

$$
\begin{aligned}
-3 \le t - 1 \le 3\\
-2 \le t \le 4\\\\
(-2, 4)
\end{aligned}
$$

```{r problem 21, echo=F, fig.height=3, fig.width=3}
p21 <- Intervals(matrix(c(-2, 4), ncol=2), closed = c(T, T))
plot(p21)
```

#### 22.

$|t + 2| < 1$

$$
\begin{aligned}
-1 < t + 2 < 1\\
-3 < t < -1\\\\
(-3, -1)
\end{aligned}
$$

```{r problem 22, echo=F, fig.height=3, fig.width=3}
p22 <- Intervals(matrix(c(-3, -1), ncol=2), closed = c(F, F))
plot(p22)
```

#### 23.

$|3y - 7| < 4$

$$
\begin{aligned}
-4 < 3y - 7 < 4\\
3 < 3y < 11\\
1 < y < \frac{11}{3}\\\\
\left(1, \frac{11}{3}\right)
\end{aligned}
$$

```{r problem 23, echo=F, fig.height=3, fig.width=3}
p23 <- Intervals(matrix(c(1, 11/3), ncol=2), closed = c(F, F))
plot(p23)
```

#### 24.

$|2y + 5| < 1$

$$
\begin{aligned}
-1 < 2y + 5 < 1\\
-6 < 2y < -4\\
-3 < y < -2\\\\
(-3, -2)
\end{aligned}
$$

```{r problem 24, echo=F, fig.height=3, fig.width=3}
p24 <- Intervals(matrix(c(-3, -2), ncol=2), closed = c(F, F))
plot(p24)
```

#### 25.

$\left|\frac{z}{5} - 1\right| \le 1$

$$
\begin{aligned}
-1 \le \frac{z}{5} - 1 \le 1\\
-5 \le z \le 5\\\\
[-5, 5]
\end{aligned}
$$

```{r problem 25, echo=F, fig.height=3, fig.width=3}
p25 <- Intervals(matrix(c(-5, 5), ncol=2), closed = c(F, F))
plot(p25)
```

#### 26.

$\left|\frac{3}{2}z - 1\right| \le 2$

$$
\begin{aligned}
-2 \le \frac{3}{2}z - 1 \le 2\\
-4 \le 3z - 2 \le 4\\
-2 \le 3z \le 6\\
-\frac{2}{3} \le z \le 2\\\\
\left[-\frac{2}{3}, 2\right]
\end{aligned}
$$

```{r problem 26, echo=F, fig.height=3, fig.width=3}
p26 <- Intervals(matrix(c(-2/3, 2), ncol=2), closed = c(T, T))
plot(p26)
```

#### 27.

$\left|3 - \frac{1}{x}\right| < \frac{1}{2}$

$$
\begin{aligned}
-\frac{1}{2} < 3 - \frac{1}{x} < \frac{1}{2}\\\\
-1 < 6 - \frac{2}{x} < 1\\\\
-7 < - \frac{2}{x} < -5\\\\
-\frac{1}{7} > -\frac{x}{2} > -\frac{1}{5}\\\\
7 < \frac{x}{2} < 5\\\\
14 < x < 10\\
10 > x > 14\\\\
(10, 14)
\end{aligned}
$$

```{r problem 27, echo=F, fig.height=3, fig.width=3}
p27 <- Intervals(matrix(c(10, 14), ncol=2), closed = c(F, F))
plot(p27)
```

#### 28.

$\left|\frac{2}{x} - 4\right| < 3$

$$
\begin{aligned}
-3 < \frac{2}{x} - 4 < 3\\\\
1 < \frac{2}{x} < 7\\\\
1 > \frac{x}{2} > \frac{1}{7}\\\\
2 > x > \frac{1}{14}\\
\frac{1}{14} < x < 2\\\\
\left(\frac{1}{14}, 2\right)
\end{aligned}
$$

```{r problem 28, echo=F, fig.height=3, fig.width=3}
p28 <- Intervals(matrix(c(1/14, 2), ncol=2), closed = c(F, F))
plot(p28)
```

#### 29.

$|2s| \ge 4$

$$
\begin{aligned}
&2s \le -4 &\ \ & 2s \ge 4&\\
&s \le -2 &\ \ & s \ge 2&\\\\
&(-\infty, -2)\cup(2, \infty)
\end{aligned}
$$

```{r problem 29, echo=F, fig.height=3, fig.width=3}
p29 <- Intervals(matrix(c(-20, 2, -2, 20), ncol=2), closed = c(T, T))
plot(p29, xlim=c(-4, 4))
```

#### 30.

$|s + 3| \ge \frac{1}{2}$

$$
\begin{aligned}
&s + 3 \le \frac{1}{2} &\ \ & -(s + 3) \le \frac{1}{2}&\\\\
&s \le -\frac{5}{2} &\ \ & s + 3 \ge -\frac{1}{2}&\\\\
&s \le -\frac{5}{2} &\ \ & s \ge -\frac{7}{2}&\\\\
&\left(-\infty, -\frac{7}{2}\right)\cup\left(\frac{5}{2}, \infty\right)
\end{aligned}
$$

```{r problem 30, echo=F, fig.height=3, fig.width=3}
p30 <- Intervals(matrix(c(-20, -5/2, -7/2, 20), ncol=2), closed = c(T, T))
plot(p30, xlim=c(-4, -4/2))
```

#### 31.

$|1 - x| > 1$

$$
\begin{aligned}
&1 - x > 1 &\ \ & -(1 - x) > 1&\\
&-x > 0 &\ \ & 1 - x > -1&\\
&x < 0 &\ \ & -x > -2&\\
&x < 0 &\ \ & x > 2\\\\
&(-\infty, 0)\cup(2, \infty)
\end{aligned}
$$


```{r problem 31, echo=F, fig.height=3, fig.width=3}
p31 <- Intervals(matrix(c(-20, 2, 0, 20), ncol=2), closed = c(F, F))
plot(p31, xlim=c(-1, 3))
```

#### 32.

$|2 - 3x| > 5$

$$
\begin{aligned}
&2 - 3x > 5 &\ \ & -(2 - 3x) > 5\\
&-3x > 3 &\ \ & 2 - 3x > -5\\
&3x < 3 &\ \ & -3x > -7\\
&x < 1 &\ \ & x > \frac{7}{3}\\\\
&\left(-\infty, 1\right)\cup\left(\frac{7}{3}, \infty\right)
\end{aligned}
$$

```{r problem 32, echo=F, fig.height=3, fig.width=3}
p32 <- Intervals(matrix(c(-20, 7/3, 1, 20), ncol=2), closed = c(F, F))
plot(p32, xlim=c(2/3, 8/3))
```

#### 33.

$\left|\frac{r + 1}{2}\right| \ge 1$

$$
\begin{aligned}
&\frac{r + 1}{2} \ge 1 &\ \ & -\frac{r + 1}{2} \ge 1\\\\
&r + 1 \ge 2 &\ \ & -(r + 1) \ge 2\\
&r \ge 1 &\ \ & r + 1 \le -2\\
&r \ge 1 &\ \ & r \le -3\\\\
&(-\infty, -3)\cup(1, \infty)
\end{aligned}
$$

```{r problem 33, echo=F, fig.height=3, fig.width=3}
p33 <- Intervals(matrix(c(-20, 1, -3, 20), ncol=2), closed = c(F, F))
plot(p33, xlim=c(-4, 2))
```

#### 34.

$\left|\frac{3r}{5} - 1\right| > \frac{2}{5}$

$$
\begin{aligned}
&\frac{3r}{5} - 1 > \frac{2}{5} &\ \ &
  -\left(\frac{3r}{5} - 1\right) > \frac{2}{5}\\\\
&\frac{3r}{5} > \frac{7}{5} &\ \ & \frac{3r}{5} < \frac{3}{5}\\\\
&3r > 7 &\ \ & 3r < 3\\\\
&r > \frac{7}{3} &\ \ & r < 1\\\\
&\left(-\infty, 1\right)\cup\left(\frac{7}{3}, \infty\right)
\end{aligned}
$$

```{r problem 34, echo=F, fig.height=3, fig.width=3}
p34 <- Intervals(matrix(c(-20, 7/3, 1, 20), ncol=2), closed = c(F, F))
plot(p34, xlim=c(0, 3))
```

# Quadratic Inequalities

Solve the inequalities in Exercises 35--42. Express the solution sets as
intervals or unions of intervals and graph them. Use the result
$\sqrt{a^2} = |a|$ as appropriate.

#### 35.

$x^2 < 2$

$$
\begin{aligned}
\sqrt{x^2} < \sqrt{2}\\
|x| < \sqrt{2}\\
-\sqrt{2} < x < \sqrt{2}
\end{aligned}
$$

$$
(-\sqrt{2}, \sqrt{2})
$$

```{r problem 35, echo=F, fig.height=3, fig.width=3}
p35 <- Intervals(matrix(c(-sqrt(2), sqrt(2)), ncol=2), closed = c(F, F))
plot(p35)
```

#### 36.

$4 \le x^2$

$$
\begin{aligned}
2 \le \sqrt{x^2}\\
|x| \ge 2\\
-2 < x < 2
\end{aligned}
$$

$$
(-2, 2)
$$

```{r problem 36, echo=F, fig.height=3, fig.width=3}
p36 <- Intervals(matrix(c(-2, 2), ncol=2), closed = c(T, T))
plot(p36)
```

#### 37.

$4 < x^2 < 9$

$$
\begin{aligned}
2 < \sqrt{x^2} < 3\\
2 < |x| < 3\\
(2 < x < 3)\cup(-2 > x > -3)
\end{aligned}
$$

$$
(-3, -2)\cup(2, 3)
$$

```{r problem 37, echo=F, fig.height=3, fig.width=3}
p37 <- Intervals(matrix(c(2, -3, 3, -2), ncol=2), closed = c(F, F))
plot(p37)
```

#### 38.

$\frac{1}{9} < x^2 < \frac{1}{4}$

$$
\begin{aligned}
\frac{1}{3} < \sqrt{x^2} < \frac{1}{2}\\
\left(\frac{1}{3} < x < \frac{1}{2}\right) \cup
  \left(-\frac{1}{3} > x > -\frac{1}{2}\right)
\end{aligned}
$$

$$
\left(\frac{1}{3}, \frac{1}{2}\right) \cup
  \left(-\frac{1}{2}, -\frac{1}{3}\right)
$$


```{r problem 38, echo=F, fig.height=3, fig.width=3}
p38 <- Intervals(matrix(c(1/3, -1/2, 1/2, -1/3), ncol=2), closed = c(F, F))
plot(p38)
```

#### 39.

$(x - 1)^2 < 4$

$$
\begin{aligned}
\sqrt{(x - 1)^2} < 2\\
|x - 1| < 2\\
-2 < x - 1 < 2\\
-1 < x < 3
\end{aligned}
$$

$$
(-1, 3)
$$

```{r problem 39, echo=F, fig.height=3, fig.width=3}
p39 <- Intervals(matrix(c(-1, 3), ncol=2), closed = c(F, F))
plot(p39)
```

#### 40.

$(x + 3)^2 < 2$

$$
\begin{aligned}
\sqrt{(x + 3)^2} < 2\\
|x + 3| < \sqrt{2}\\
-\sqrt{2} < x + 3 < \sqrt{2}\\
-\sqrt{2} - 3 < x < \sqrt{2} - 3
\end{aligned}
$$

$$
(-\sqrt{2} - 3, \sqrt{2} - 3)
$$

```{r problem 40, echo=F, fig.height=3, fig.width=3}
p40 <- Intervals(matrix(c(-sqrt(2)-3, sqrt(2)-3), ncol=2), closed = c(F, F))
plot(p40)
```

#### 41.

$x^2 - x < 0$

$$
x(x - 1) < 0
$$

Zeros occur at $x = 0$ and $x = 1$. This divides the real number line into the
following intervals:

$(-\infty, 0), (0, 1), (1, \infty)$

Now we test each interval

1. $(-\infty, 0)$: If $x = -1: (-1)^2 - (-1) = 2 \not< 0$

2. $(0, 1)$: If $x = .5: (.5)^2 - (.5) = -.25 < 0$ This interval works.

3. $(1, \infty)$: If $x = 1.5: (1.5)^2 - (1.5) = .75 \not< 0$

$$
(0,1)
$$

```{r problem 41, echo=F, fig.height=3, fig.width=3}
p41 <- Intervals(matrix(c(0,1), ncol=2), closed = c(F, F))
plot(p41)
```

#### 42.

$x^2 - x - 2 \ge 0$

$$
(x - 2)(x + 1) \ge 0
$$

Intervals: $(-\infty, -1], [-1, 2], [2, \infty)$

1. $(-\infty, -1)$: If $x = -1.5: (-1.5)^2 - (-1.5) - 2 = 1.75 \ge 0$ This
   interval works.

2. $(-2, 1)$: If $x = 0: 0^2 - 0 - 2 = -2 \not\ge 0$

3. $(1, \infty)$: If $x = 1.5: 1.5^2 - 1.5 - 2 = -1.25 \not\ge 0$

$$
(-\infty, -1]
$$

```{r problem 42, echo=F, fig.height=3, fig.width=3}
p42 <- Intervals(matrix(c(-20,-1), ncol=2), closed = c(F, F))
plot(p42, xlim=c(-2, 0))
```

# Theory and Examples

#### 43.

Do not fall into the trap $|-a| = a$. For what real numbers $a$ is this equation
true? For what real numbers is it false?

> This fails when $a < 0$. For all negative real numbers, the equality fails.

#### 44.

Solve the equation $|x - 1| = 1 - x$.

$$
\begin{aligned}
|x - 1| = \begin{cases}
  x - 1, x \ge 1\\
  1 - x, x < 1
\end{cases}
\end{aligned}
$$

for case 1: $x \ge 1$

$$
\begin{aligned}
x - 1 = 1 - x\\
2x = 2\\
x = 1
\end{aligned}
$$

For case 2: $x < 1$

$$
x - 1 = x - 1\\
$$

This is an __identity__ (true for all $x < 1$)

$$
(-\infty, 1]
$$

```{r problem 44, echo=F, fig.height=3, fig.width=3}
p44 <- Intervals(matrix(c(-20,1), ncol=2), closed = c(F, T))
plot(p44, xlim=c(-2, 1))
```

#### 45. A proof of the triangle inequality.

Give the reason justifying each of the numbered steps in the following proof of
the triangle inequality.

$$
\begin{aligned}
|a + b|^2 &= (a + b)^2\\
  &= a^2 + 2ab + b^2\\
  &\le a^2 + 2|a||b| + b^2\\
  &\le |a|^2 + 2|a||b| + |b|^2\\
  &= (|a| + |b|)^2\\
|a + b| &\le |a| + |b|
\end{aligned}
$$

#### 46.

Prove that $|ab| = |a||b|$ for any number $a$ and $b$.

$$
\begin{aligned}
|ab| &= \sqrt{(ab)^2}\\
  &= \sqrt{a^2} \times \sqrt{b^2}\\
  &= |a||b|
\end{aligned}
$$

#### 47.

If $|x| \le 3$ and $x > -1/2$, what can you say about $x$?

#### 48.

Graph the inequality $|x| + |y| \le 1$.

#### 49. GRAPHER

##### a.

Graph the functions $f(x) = x/2$ and $g(x) = 1 + (4/x)$ together to identify the
values of $x$ for which

$$
\frac{x}{2} > 1 + \frac{4}{x}.
$$

##### b.

Confirm your findings in (a) algebraically.

#### 50. GRAPHER

##### a.

Graph the functions $f(x) = 3/(x - 1)$ and $g(x) = 2/(x + 1)$ together to
identify the values of $x$ for which

$$
\frac{3}{x - 1} < \frac{2}{x + 1}.
$$

##### b.

Confirm your findings in (a) algebraically.
