---
title: Preliminaries
subtitle: 1 Real Numbers and the Real Line
output: pdf_document
toc: true
header-includes:
  - \usepackage{tcolorbox}
  - \input{../helpers/latex.tex}
---

```{r setup, include=F}
options(repos=c("https://mirrors.nics.utk.edu/cran/"))
suppressMessages({
    if(!require(tidyverse)){
        install.packages("tidyverse")
        require(tidyverse)
    }
    if(!require(intervals)){
        install.packages('intervals')
        require(intervals)
    }
    if(!require(knitr)){
        install.package('knitr')
        require(knitr)
    }
})
source('../helpers/graph_functions.R')
```

# Real Numbers and the Real Line

Much of calculus is based on properties of the real number system. __Real
numbers__ are numbers that can be expressed as decimals, such as:

$$
\begin{aligned}
-\frac{3}{4} &= -0.75000\ldots\\\\
\frac{1}{3} &= 0.33333\ldots\\\\
\sqrt{2} &= 1.4142\ldots\\\\
\end{aligned}
$$

The dots $\ldots$ in each case indicate that the sequence of decimal digits goes
on forever.

The real numbers can be represented geometrically as points on a number line
called the __real line__.

```{r real number line, echo=F, fig.height=2}
c0.1 <- new.env()
c0.1$df <- tibble(
    x = c(-2, -1, -3/4, 0, 1/3, 1, sqrt(2), 2, 3, pi, 4),
    y = rep(0, 11),
    labels = c("-2", "-1", "-frac(3,4)", "0", "frac(1,3)", "1", "sqrt(2)",
               "2", "3", "pi", "4")
)

ggplot(c0.1$df, aes(x, y)) +
    geom_hline(yintercept = 0) +
    #geom_segment(aes(xend = x, y = -0.05, yend = 0.05)) +
    geom_text(aes(label = labels), vjust = 2.25, parse=T, size=2.8) +
    geom_text(aes(label = rep('|', length(x)))) +
    scale_y_continuous(breaks = NULL) +
    xlab("") +
    ylab("") +
    theme(
        panel.grid.major.x = element_blank(),
        panel.grid.minor.x = element_blank(),
        axis.text.x = element_blank(),
        axis.ticks.x = element_blank()
    )
        
```

The symbol $\mathbb{R}$ denotes either the real number system or, equivalently,
the real line.

# Properties of Real Numbers

The properties of the real number system fall into three categories: algebraic
properties, order properties, and completeness. The algebraic properties say
that the real numbers can be added, subtracted, multiplied, and divided (except
by 0) to produce more real numbers under the usual rules of arithmetic.
_you can never divide by 0_.

The order properties of real numbers are summarized in the following list.

\begin{bluebox}

\subsection{Rules for Inequalities}

If $a, b,$ and $c$ are real numbers, then:

1. $a < b \Rightarrow a + c < b + c$

2. $a < b \Rightarrow a - c < b - c$

3. $a < b \text{ and } c > 0 \Rightarrow ac < bc$

4. $a < b \text{ and } c < 0 \Rightarrow bc < ac$
   (Special case: $a < b \Rightarrow -b < -a$

5. $a > 0 \Rightarrow \frac{1}{a} > 0$

6. If $a$ and $b$ are both positive or both negative, then
   $a < b \Rightarrow \frac{1}{b} < \frac{1}{a}$

\end{bluebox}

The symbol $\Rightarrow$ means "implies".

Notice the rules for multiplying an inequality by a number. Multiplying by a
positive number preserves the inequality; multiplying by a negative number
reverses the inquality. Also, reciprocation reverses the inequality for numbers
of the same sign.

The completeness property of the real number system is deeper and harder to
define precisely. Roughly speaking, it says that there are enough real numbers
to "complete" the real number line, in the sense that there are no "holes" or
"gaps" in it. Many of the theorems of calculus would fail if the real number
system were not complete, and the nature of the connection is important. The
topic is best saved for a more advanced course, however, and we will not pursue
it.

# Subsets of $\mathbb{R}$

We distinguish three special subsets of real numbers.

1. The __natural numbers__, namely $1, 2, 3, 4, \ldots$

2. The __integers__, namely $0, \pm1, \pm2, \pm3, \pm4, \ldots$

3. The __rational numbers__, namely the numbers that can be expressed in the
   form of a fraction $m/n$, where $m$ and $n$ are integers and $n \ne 0$.
   Examples are

$$
\frac{1}{3}, -\frac{4}{9}, \frac{200}{13}, \text{ and } 57 = \frac{57}{1}.
$$

The rational numbers are precisely the real numbers with decimal expansions that
are either

a. terminating (ending in an infinite string of zeros), for example,

$$
\frac{3}{4} = 0.75000\ldots = 0.75 \text{ or }
$$

b. repeating (ending with a block of digits that repeats over and over), for
example

$$
\frac{23}{11} = 2.090909\ldots = 2.\overline{09}
$$

> The bar indicates the block of repeating digits.

The set of rational numbers has all the algebraic and order properties of the
real numbers but lacks the completeness property. For example, there is no
rational number whose square is 2; there is a "hole" in the rational line where
$\sqrt{2}$ should be.

Real numbers are not rational are called __irrational numbers__. They are
characterized by having nondetermining and non repeating deciam expansions.
Examples are $\pi$, $\sqrt{2}$, $\sqrt[3]{5}$, and $\log_{10}{3}$.

# Intervals

A subset of the real line is called an __interval__ if it contains at least two
numbers and contains all the real numbers lying between any two of its elements.
For example, the set of all real numbers $x$ such that $x > 6$ is an interval,
as is the set of all $x$ such taht $-2 \le x \le 5$. The set of all nonzero real
numbers is not an interval; since 0 is absent, the set fails to contain every
real number between $-1$ and $1$ (for example).

Geometrically, intervals of numbers corresponding to line segments are
__finite intervals__.

A finite interval is said to be __closed__ if it contains both of its endpoints,
__half-open__ if it contians one endpoint but not the other, and __open__ if it
contains neither endpoint. THe endpoints are also called __boundary points__;
they make up the interval's __boundary__. The remaining points of the interval
are __interior points__ and together make up what is called the interval's
__interior__.


```{r table 1, echo=F}
p <- line.graph(Intervals(matrix(c(0, 5), ncol=2), closed=c(F,F)))
```

# Solving Inequalities

The process of finding the interval or intervals of numbers that satisfy an
inequality in $x$ is called __solving__ the inequality.

__EXAMPLE 1__
Solve the following inequalities and graph their solution sets on the real line.

a. $2x - 1 < x + 3$

b. $-\frac{x}{3} < 2x + 1$

c. $\frac{6}{x-1} \ge 5$

__Solution__

a.

$$
\begin{aligned}
2x - 1 &< x + 3\\
2x &< x + 4\\
x &< 4
\end{aligned}
$$

The solution set is the interval $(-\infty, 4)$ (Fig. 1a).

b.

$$
\begin{aligned}
-\frac{x}{3} &< 2x + 1\\
-x &< 6x + 3\\
0 &< 7x + 3\\
-3 &< 7x\\
-\frac{3}{7} &< x
\end{aligned}
$$

The solution set is the interval $(-3/7, \infty)$ (Fig. 1b).

c.

The inequality $6/(x - 1) \ge 5$ can hold only if $x > 1$, because otherwise
$6/(x - 1)$ is undefined or negative. Therefore, the inequality will be
preserved if we multiply both sides by $(x - 1)$, and we have

$$
\begin{aligned}
\frac{6}{x-1} &\ge 5\\
6 &\ge 5x - 5\\
11 &\ge 5x\\
\frac{11}{5} &\ge x.
\end{aligned}
$$

The solution set is the half-open interval $(1, 11/5]$ (Fig. 1c).

# Absolute Value

The __absolute value__ of a number $x$, denoted by $|x|$, is defined by the
formula

$$
\begin{aligned}
|x| =
\begin{cases}
  x, & x \ge 0 \\
  -x, & x < 0.
\end{cases}
\end{aligned}
$$

__EXAMPLE 2__
$|3| = 3$, $|0| = 0$, $|-5| = -(-5) = 5$, $|-|a|| = |a|$

Notice that $|x| \ge 0$ for every real number $x$, and $|x| = 0$ if and only if
$x = 0$.

Since the symbol $\sqrt{a}$ always denotes the _nonnegative_ square root of $a$,
an alternate definition of $|x|$ is

$$
|x| = \sqrt{x^2}
$$

\begin{bluebox}
It is important to remember that $\sqrt{a^2} = |a|$. Do not write
$\sqrt{a^2} = a$ unless you already know that $a \ge 0$.
\end{bluebox}

\begin{bluebox}

\begin{tabular}{lll}

\multicolumn{3}{l}{\textbf{Absolute Value Properties}}\\

\vspace{.1in}

\textbf{1.} & $|-a| = |a|$
  & \parbox{10cm}{A number and its negative have the same absolute
  value.}\\

\vspace{.1in}

\textbf{2.} & $|ab| = |a||b|$
  & \parbox{10cm}{The absolute value of a product is the product of
    the absolute values.}\\

\vspace{.1in}

\textbf{3.} & $\left|\frac{a}{b}\right| = \frac{|a|}{|b|}$
  & \parbox{10cm}{The absolute value of a quotient is the quotient of the
    absolute values.}\\

\textbf{4.} & $|a + b| \le |a| + |b|$
  & \parbox{10cm}{\textbf{The triangle inequality} The absolute value of the sum
    of two numbers is less than or equal to the sum of their absolute values.}\\

\end{tabular}

\end{bluebox}

If $a$ and $b$ differ in sign, then $|a + b|$ is less than $|a| + |b|$. In all
other cases, $|a + b|$ equals $|a| + |b|$.

__EXAMPLE 3__

$$
\begin{aligned}
|-3 + 5| = |2| < |-3| + |5| = 8\\
|3 + 5| = |8| = |3| + |5|\\
|-3 - 5| = |-8| = 8 = |-3| + |-5|
\end{aligned}
$$

__EXAMPLE 4__ Solve the equation $|2x - 3| = 7$.

__Solution__ The equation says that $2x - 3 = \pm7$, so there are two
possibilities:

$$
\begin{aligned}
2x - 3 &= 7 & 2x - 3 &= -7\\
2x &= 100   & 2x &= -4\\
x &= 5 & x &= -2
\end{aligned}
$$

## Inequalities Involving Absolute Values

The inequality $|a| < D$ says that the distance from $a$ to $0$ is less than
$D$. Therefore, $a$ must lie between $D$ and $-D$.

\begin{bluebox}

\subsubsection{Intervals and Absolute Values}

If D is any positive number, then

$$
\begin{aligned}
|a| < D & \Leftrightarrow & -D < a < D,
\end{aligned}
$$

$$
\begin{aligned}
|a| < D & \Leftrightarrow & -D \le a \le D.
\end{aligned}
$$

\end{bluebox}

__EXAMPLE 5__ Solve the inequality $|x - 5| < 9$ and graph the solution set on
the ral line.

__Solution__

$$
\begin{aligned}
|x - 5| < 9\\
-9 < x - 5 < 9\\
-9 + 5 < x < 9 + 5\\
-4 < x < 14
\end{aligned}
$$

The solution set is the open interval (-4, 14) (Fig 3.)

```{r figure 3, echo=F}
line.graph(Intervals(matrix(c(-4, 14), ncol=2), closed = c(F, F)))
```

__EXAMPLE 6__ Solve the inequality $\left|5 - \frac{2}{x}\right| < 1$.

__Solution__ We have

$$
\begin{aligned}
\left|5 - \frac{2}{x}\right| < 1 &\Leftrightarrow -1 < 5 - \frac{2}{x} < 1\\
  &\Leftrightarrow -6 < -\frac{2}{x} < -4\\
  &\Leftrightarrow 3 > \frac{1}{x} > 2\\
  &\Leftrightarrow \frac{1}{3} < x < \frac{1}{2}.
\end{aligned}
$$

Notice how the vairous rules for inequalities were used here. Multiplying by a
negative number reverses the inequality. So does taking reciprocals in an
inequality in which both sides are positive. The original inequality holds if
and only if $(1/3) < x < (1/2)$. The solution set is the open interval
$(1/3, 1/2)$.

```{r example 2 figure, echo=F}
line.graph(Intervals(matrix(c(1/3, 1/2), ncol=2), closed = c(F, F)))
```

__EXAMPLE 7__ Solve the inequality and graph the solution set:

a. $|2x - 3| \le 1$

b. $|2x - 3| \ge 1$

__Solution__

a.

$$
\begin{aligned}
|2x - 3| \le 1\\
-1 \le 2x - 3 \le 1\\
2 \le 2x \le 4\\
1 \le x \le 2
\end{aligned}
$$

b.

$$
\begin{aligned}
|2x - 3| &\ge 1\\
2x - 3 \ge 1\ \ &\text{or}\ \ -(2x - 3) \ge 1\\
2x - 3 \ge 1\ \ &\text{or}\ \ 2x - 3 \le -1\\
x - \frac{3}{2} \ge \frac{1}{2}\ \ &\text{or}\ \
  x - \frac{3}{2} \le -\frac{1}{2}\\
x \ge 2\ \ &\text{or}\ \ x \le 1
\end{aligned}
$$

The solution set is $(-\infty, 1] \cup [2, \infty)$ (fig. 4b).

```{r fig 4, echo=F}
line.graph(Intervals(matrix(c(1, 2), ncol=2), closed=c(T, T)))
line.graph(Intervals(matrix(c(-Inf, 1), ncol=2), closed=c(T, T)))
```

## Union and intersection

Notice the use of the symbol $\cup$ to denote the union of intervals. A number
lies in the __union__ of two sets if it lies in either set. Similarly we use the
symbol $\cap$ to denote intersection. A number lies in the intersection
$I \cap J$ of two sets if it lies in _both_ sets $I$ and $J$. For example,
$[1, 3)\cap[2, 4] = [2, 3)$.
