---
title: Preliminaries
subtitle: Real Numbers and the Real Line
output:
  pdf_document:
    toc: true
    includes:
      in_header: "../preamble.tex"
---

---

# Error patterns to focus on

## Confusing "necessarily true" with "identical," and testing coutnerexamples from the wrong set.

(Task 7, Problem 4 parts c/d)

You initially checked whether two intervals were _equal_ rather than whether the
given interval was a _subset_ of the candidate's interval. Related to this, when
looking for counterexamples, you sometimes picked test values _outside_ the
given interval, which can never disprove necessity.

_Fix_: Before testing, fix the rule: a counterexample must always come from
__within the given condition__. Ask "is every element of the given set also in
the candidate set?"--never check for exact equality, and never test a point
outside the assumption.

## Bracket/strictness mismatches between the inequality and the interval notation

(#6, #36, #47 twice)

Non-strict inequalities ($\le$, $\ge$) need closed brackets; strict ones ($<$,
$>$) need open brackets $-$ and $\pm\infty$ always get a parenthesis, never a
bracket, regardless of the inequality. When merging two conditions with
_different_strictness (like #47), each surviving endpoint keeps the strictness
of whichever single condition produced it.

_Fix_: After finishing the algebra, do a dedicated final pass just checking each
interval endpoint against its orginating inequality--treat it as a separate step
from solving.

## Sign-dropping in absolute value case splits

(#42 first attempt, part of #47)

When splitting $|A| \ge c$ or $|A| \le c$ into two cases, a $-c$ occasionally
got written as $+c$.

_Fix_: Always write the case template explicitly before substituting: $|A| \ge c
\Leftrightarrow A \ge c \land A \le -c$. Don't skip straight to numbers.

## Product/quotient sign-rule errors

(#41 first attempt)

Treat "$AB < 0$" as "both factors negative" instead of "factors have opposite
signs."

_Fix_: Memorize: same signs $\rightarrow$ positive product; opposite signs
$\rightarrow$ negative product. Always write out both opposite-sign cases
explicitly rather than reasoning informally.

## Logical connective misuse ($\land$ vs. $\lor$, $\Leftrightarrow$ vs. $=$)

(#44, #46)

Used AND where OR was needed (absolute value equation cases), and used
$\Leftrightarrow$ where a plain equality chain was intended.

_Fix_: Before writing a connective, ask "am I stating two things that are equal,
or one thing implying another, or a case split where only one needs to hold?"
Equalities get $=$; case splits from $|A| = c$ get $\lor$; only use
$\Rightarrow$ when one statement's truth forces another's.

## Multiplying/dividing an inequality by an expression of unknown sign

(#49, #50)-- Your most consequential and recurring error type.

Cleared denominators or took reciprocals without first splitting into sign
cases, silently assuming positivity and losing solution branches.

_Fix_: Whenever a variable expression multiplies or divides both sides of an
inequality, __stop and ask: could this expression be negative?__ If yes, split
into cases _before_ clearing denominators--never combine the case-split with an
already-cleared equation (this caused the leftover unconditional line in your
first #49 revision).

## Boundary/edge cases treated loosely

(#43 initially excluded $a = 0$)

_Fix_: When a problem asks "for which values does X hold," explicitly test
boundary points (0, endpoints, asymptotes) rather than just describing open
regions.

## Graphing-code domain bugs

(#49 missing $+1$ term, #50 domain not spanning all solution branches or
vertical asymptotes)

_Fix_: Before writing `seq()`/`plot()` calls, identify vertical
asymptotes/undefined points first, and split the domain into separate segments
around them--this should be a fixed first step, not an afterthought.

## The single biggest theme to focus on

Operations that are only valid under a hidden assumption (multiplying by a
possibly-negative quantity, taking reciprocals, matching bracket strictness,
picking counterexamples from the right set). In nearly every error, the fix was
the same discipline--__pause before an algebraic shortcut and ask what condition
makes it valid, then either verify that condition holds or split into cases.__
This is worth making a standing habit, since it accounts for the majority of the
substantive (non-typo) mistakes above.

# Summary of Topics

This section reviews real numbers, inequalities, intervals, and absolute values.


## Real Numbers and the Real Line

Much of calculus is based on properties of the real number system. __Real
numbers__ are numbers that can be expressed as decimals, such as

$$
\begin{aligned}
-\frac{3}{4} &= -0.75000...\\
\frac{1}{3}  &=  0.33333...\\
\sqrt{2}     &=  1.4142...\\
\end{aligned}
$$

The dots $...$ in each case indicate that the sequence of decimal digits goes on forever. 

The real numbers can be represented geometrically as points on a number line
called the __real line__.

```{r real-line, echo=F}
points <- c(-2, -1, -3/4, 0, 1/3, 1, sqrt(2), 2, 3, pi, 4)
labels <- c("-2", "-1", "-3/4", "0", "1/3", "1,", expression(sqrt(2)), "2", "3",
            expression(pi), "4")

plot(points, rep(0, length(points)), type = "n",
     xlim = c(-2.5, 4.5), ylim = c(-1, 1),
     axes = FALSE, xlab = "", ylab = "")

abline(h=0)
points(points, rep(0, length(points)), pch = "|")
text(points, rep(-0.2, length(points)), labels = labels, cex = 0.7)
```

The symbol $\mathbb{R}$ donotes either the real number system or, equivalently,
the real line.

## Properties of Real Numbers

The properties of the real number system fall into three categories: algebraic
properties, order properties, and completeness. The algebraic properties say
that the real numbers can be added, subtracted, multiplied, and divided (except
by 0) to produce more real numbers under the usual rules of arithmetic. _You can
never divide by 0_.

The order properties of real numbers are summarized in the following list.

\begin{tcolorbox}
\textbf{Rules for Inequalities}

If $a$, $b$, and $c$ are real numbers, then:

\begin{enumerate}
\item $a < b \Rightarrow a + c < b + c$
\item $a < b \Rightarrow a - c < b - c$
\item $a < b$ and $c > 0 \Rightarrow ac < bc$
\item $a < b$ and $c < 0 \Rightarrow bc < ac$\newline
      Special case: $a < b \Rightarrow -b < -a$
\item $a > 0 \Rightarrow \frac{1}{a} > 0$
\item If $a$ and $b$ are both positive or both negative, then $a < b \Rightarrow
\frac{1}{b} < \frac{1}{a}$.
\end{enumerate}
\end{tcolorbox}

The completeness property of the real number system is explored in Rudin's
_"Principles of Mathematical Analysis"_. It is ddper and harder to define
precisely. Roughly speaking, it says that there are enough real numbers to
"complete" the real number line, in the sense that there are no "holes" or
"gaps" in it. Many of the theorems of calculus would fail if the real number
system were not complete, and the nature of the connection is important.

## Subsets of $\mathbb{R}$

We distinguish three special subsets of real numbers.

1. The natural numbers: $1, 2, 3, 4, ...$
2. The integers: $0, \pm1, \pm2, \pm3, ...$
3. The rational numbers. Numbers that can be expressed in the form of a fraction
   $m/n$, where $m$ and $n$ are integers and $n \ne 0$:

$$
\frac{1}{3}, -\frac{4}{9}, \frac{200}{13}, \text{and} 57 = \frac{57}{1}.
$$

> The rational numbers are precisely the real numbers with decimal expansions
  that either
> a) terminating (ending in an infinite string of zeros):

$$
\frac{3}{4} = 0.750000... = 0.75 \quad \text{or}
$$

> b) repeating (ending with a block of digits that repeats over and over):

$$
\frac{23}{11} = 2.090909... = 2.\overline{09}
$$

The set of rational numbers has all the algebraic and order properties of the
real numbers but lacks the completeness property. For example, there is no
rational number whose square is 2; there is a "hole" in the rational line where
$\sqrt{2}$ should be.

Real numbers that are not rational are called __irrational numbers__. They are
characterized by having nonterminating and nonrepeating decimal expansions:
$\pi$, $\sqrt{2}$, $\sqrt[3]{5}$, and $\log_{10}{3}$.

## Intervals

\begin{tcolorbox}
\begin{itemize}
\item $(a, b) \quad \{x \mid a < x < b\} \quad$
    \intervalgraph{open}{open}{a}{b}
\item $[a, b] \quad \{x \mid a \le x \le b\} \quad$
    \intervalgraph{closed}{closed}{a}{b}
\item $[a, b) \quad \{x \mid a \le x < b\} \quad$
    \intervalgraph{closed}{open}{a}{b}
\item $(a, b] \quad \{x \mid a < x \le b\} \quad$
    \intervalgraph{open}{closed}{a}{b}
\item $(a, \infty) \quad \{x \mid x > a\} \quad$
    \intervalgraph{open}{inf}{a}{}
\item $[a, \infty) \quad \{x \mid x \ge a\} \quad$
    \intervalgraph{closed}{inf}{a}{}
\item $(-\infty, b) \quad \{x \mid x < b\} \quad$
    \intervalgraph{inf}{open}{}{b}
\item $(-\infty, b] \quad \{x \mid x \le b\} \quad$
    \intervalgraph{inf}{closed}{}{b}
\item $(-\infty, \infty) \quad \mathbb{R}$ (set of all real numbers)
    $\quad$ \intervalgraph{inf}{inf}{}{}
\end{itemize}
\end{tcolorbox}

## Solving Inequalities

### Example 1

Solve the following inequalities and graph their solution sets on the real line.

\begin{proof}
Solutions
a) $2x - 1 < x + 3$
> \begin{solution}
$$
\begin{aligned}
2x - 1 &< x + 3\\
2x &< x + 4\\
x &< 4
\end{aligned}
$$
The solution set is the interval $(-\infty, 4)  \quad$
\intervalgraph{inf}{open}{}{4}
\end{solution}
b) $-\frac{x}{3} < 2x + 1$
> \begin{solution}
$$
\begin{aligned}
-\frac{x}{3} &< 2x + 1\\
-x &< 6x + 3\\
0  &< 7x + 3\\
-3 &< 7x\\
-\frac{3}{7} &< x
\end{aligned}
$$
The solution set is the interval $(-3/7, \infty) \quad$
\intervalgraph{open}{inf}{-3/7}{}
\end{solution}
c) $\frac{6}{x - 1} \ge 5$
> \begin{solution}
The inequality $6/(x - 1) \ge 5$ can hold only if $x > 1$, becuase otherwise
$6/(x - 1)$ is undefined or negative. Therefore, the inequality will be
preserved if we multiply both sides by $(x - 1)$.
$$
\begin{aligned}
\frac{6}{x - 1} &\ge 5\\
6 &\ge 5x - 5\\
11 &\ge 5x\\
\frac{11}{5} &\ge x
\end{aligned}
$$
The solution set is the half-open interval $(1, 11/5] \quad$
\intervalgraph{open}{closed}{1}{11/5}
\end{solution}
\end{proof}

## Absolute Value

$$
\begin{aligned}
|x| = \begin{cases}&\quad x, \quad x \ge 0\\&-x, \quad x < 0.\end{cases}
\end{aligned}
$$

### Example 2

\begin{proof}
$|3| = 3, \quad |0| = 0, \quad |-5| = -(-5) = 5, \quad |-|a|| = |a|$
\end{proof}

Notice that $|x| \ge 0$ for every real number $x$, and $|x| = 0$ if and only if
$x = 0$.

Since the symbol $\sqrt{a}$ always denotes the _nonnegative_ square root of $a$,
an alternate definition of $|x|$ is

$$
|x| = \sqrt{x^2}
$$

Geometrically, $|x|$ represents the distance from $x$ to the origin $0$ on the
real line. More generally

$$
|x - y| = \text{ the distance between x and y}
$$

\begin{tcolorbox}
\textbf{Absolute Value Properties}
\begin{enumerate}
\item $|-a| = |a| \quad$ A number and its negative have the same absolute
value.\newline
\item $|ab| = |a||b| \quad$ The absolute value of a product is the product of
the absolute values.\newline
\item $\left|\frac{a}{b}\right| = \frac{|a|}{|b|} \quad$ The absolute value of a
quotient is the quotient of the absolute values.\newline
\item $|a + b| \le |a| + |b| \quad$ \textbf{The triangle inequality} The
absolute value of the sum of two numbers is less than or equal to the sum of
their
absolute values.\newline
\end{enumerate}
\end{tcolorbox}

### Example 3

\begin{proof}
$$
\begin{aligned}
|-3 + 5| &= |2| = 2 < |-3| + |5| = 8\\
|3 + 5| &= |8| = |3| + |5|\\
|-3 - 5| &= |-8| = 8 = |-3| + |-5|
\end{aligned}
$$
\end{proof}

### Example 4

Solve the equation $|2x - 3| = 7$.

\begin{proof}
The equation says that $2x - 3 = \pm7$, so there are two possibilities:
\begin{solution}
$$
\begin{aligned}
2x - 3 &= 7 & 2x - 3 &= -7\\
2x &= 10 & 2x &= -4\\
x &= 5 & x &= -2
\end{aligned}
$$
The solutions of $|2x - 3| = 7$ are $x = 5$ and $x = -2$.
\end{solution}
\end{proof}

## Inequalities Involving Absolute Values

\begin{tcolorbox}
If $D$ is any positive number, then
\begin{equation}
\begin{aligned}
|a| &< D \quad &\Leftrightarrow \quad -D < a < D\\
\end{aligned}
\end{equation}
\begin{equation}
\begin{aligned}
|b| &\le D \quad &\Leftrightarrow \quad -D \le a \le D
\end{aligned}
\end{equation}
\end{tcolorbox}

### Example 5

Solve the inequality $|x - 5| < 9$ and grpah the solution set on the real line.

\begin{solution}
$$
\begin{aligned}
|x - 5| &< 9\\
-9 < x - 5 &< 9\\
-9 + 5 < x &< 9 + 5\\
-4 < x &< 14
\end{aligned}
$$
The solution set is the open interval $(-4, 14)$.
\intervalgraph{open}{open}{-4}{14}
\end{solution}

### Example 6

Solve the inequality $\left|5 - \frac{2}{x}\right| < 1$.

\begin{solution}
We have
$$
\begin{aligned}
\left|5 - \frac{2}{x} \right| < 1 &\Leftrightarrow -1 < 5 - \frac{2}{x} < 1\\
\quad &\Leftrightarrow -6 < -\frac{2}{x} < -4\\
\quad &\Leftrightarrow 3 > \frac{1}{x} > 2\\
\quad &\Leftrightarrow \frac{1}{3} < x < \frac{1}{2}
\end{aligned}
$$
Notice how the various rules for inequalities were used here. Multiplying by a
negative number reverses the inequality. So does taking reciprocals in an
inequality in which both sides are positive. The original inequality holds if
and only if $(1/3) < x < (1/2)$. The solution set is the open interval $(1/3,
1/2)$.
\end{solution}

### Example 7

Solve the inequality and graph the solution set

\begin{multicols}{2}
\begin{enumerate}[label=\textbf{\alph*})]
\item $|2x - 3| \le 1$
\item $|2x - 3| \ge 1$
\end{enumerate}
\end{multicols}

\begin{solution}
\begin{enumerate}[label=\textbf{\alph*})]
\item
$$
\begin{aligned}
|2x - 3| &\le 1\\
-1 \le 2x - 3 &\le 1\\
2 \le 2x &\le 4\\
1 \le x &\le 2
\end{aligned}
$$
The solution set is the closed interval $[1,2]$
\intervalgraph{closed}{closed}{1}{2}
\item
$$
\begin{aligned}
|2x - 3| \ge 1\\
2x - 3 &\ge 1 \quad \text{or} \quad &-(2x - 3) &\ge 1\\
2x - 3 &\ge 1 \quad \text{or} \quad &2x - 3 &\le -1\\
x - \frac{3}{2} &\ge \frac{1}{2} \quad \text{or}
    \quad &x - \frac{3}{2} &\le - \frac{1}{2}\\
x &\ge 2 \quad \text{or} \quad &x &\le 1
\end{aligned}
$$
\end{enumerate}
The solution set is $(-\infty, 1]\cup[2, \infty)$
\intervalgraph{inf}{closed}{}{1} \intervalgraph{closed}{inf}{2}{}
\end{solution}

# Exercises

## Decimal Representations

1. Express 1/9 as a repeating decimal, using a bar to indicate the repeating
   digits. What are the decimal representations of 2/9? 3/9? 8/9?

> \begin{answer}
$$
\begin{aligned}
\overline{111}\\
\overline{333}\\
\overline{888}
\end{aligned}
$$
\end{answer}

---

2. Express 1/11 as a repeating decimal, using a bar to indicate the repeating
   digits. What are the decimal representations of 2/11? 3/11? 9/11?

> \begin{answer}
$$
\begin{aligned}
0.\overline{090909}\\
0.\overline{181818}\\
0.\overline{272727}\\
0.\overline{818181}
\end{aligned}
$$
\end{answer}

## Inequalities

3. If $2 < x < 6$, which of the following statements about $x$ are necessarily
   true, and which are not necessarily true?

> \begin{tasks}(2)
\task $0 < x < 4$
\task $0 < x - 2 < 4$
\task $1 < \frac{x}{2} < 3$
\task $\frac{1}{6} < \frac{1}{x} < \frac{1}{2}$
\task $1 < \frac{6}{x} < 3$
\task $|x - 4| < 2$
\task $-6 < -x < 2$
\task $-6 < -x < -2$
\end{tasks}

> \begin{answer} Assume $2 < x < 6$
\begin{enumerate}[label=\alph*)]
\item Not necessarily true. Let $x = 5$; this satisfies $2 < x < 6$, but $5
\ngtr 4$, so $0 < x < 4$ fails.
\item True. Add 2 to all sides.
\item True. Multiply all terms by 2.
\item True. Its the inverse of the statement.
$$
\begin{aligned}
&\frac{1}{6} < \frac{1}{x} < \frac{1}{2}\\
&\frac{1}{6} < \frac{1}{x} \land \frac{1}{x} < \frac{1}{2}\\
&\frac{1}{6} < \frac{1}{x} \Leftrightarrow 6 > x\\
&\frac{1}{x} < \frac{1}{2} \Leftrightarrow x > 2\\
&6 > x \land x > 2 \Leftrightarrow 2 < x < 6
\end{aligned}
$$
\item True. Divide by 6 to get the inverse of the statement.
\item True.
    $|x - 4| < 2 \Rightarrow -2 < x - 4 < 2 \Rightarrow 2 < x < 6$.
\item Necessarily True. $-6 < -x < 2 \Rightarrow -2 < x < 6$.
\item True. $-6 < -x < -2 \Rightarrow 2 < x < 6$.
\end{enumerate}
\end{answer}

---

4. If $-1 < y - 5 < 1$, which of the following statements about $y$ are
   necessarily true, and which are not necessarily true?

> \begin{tasks}(2)
\task $4 < y < 6$
\task $-6 < y < -4$
\task $y > 4$
\task $y < 6$
\task $0 < y - 4 < 2$
\task $2 < \frac{y}{2} < 3$
\task $\frac{1}{6} < \frac{1}{y} < \frac{1}{4}$
\task $|y - 5| < 1$
\end{tasks}

> \begin{answer} Assume $-1 < y - 5 < 1 \Rightarrow 4 < y < 6$
\begin{enumerate}[label=\alph*)]
\item True. This is the same interval.
\item Not True. $y < -4 \Rightarrow y < 4$.
\item True. $y > 4$ is directly implied by $4 < y < 6$.
\item True. $y < 6$ is directly implied by $4 < y < 6$.
\item True. $0 < y - 4 < 2 \Rightarrow 4 < y < 6$.
\item True. $2 < \frac{y}{2} < 3 \Rightarrow 4 < y < 6$.
\item True. Its the inverse of the statement (see 3b for deriviation of logic).
\item True. $|y - 5| < 1 \Rightarrow -1 < y - 5 < 1 \Rightarrow 4 < y < 6$.
\end{enumerate}
\end{answer}

In Exercises 5--12, solve the inequalities and graph the solution sets.

---

5. $-2x > 4$

> \begin{answer}
$$
\begin{aligned}
-2x &> 4\\
x &< -2
\end{aligned}
$$
The interval is $(-\infty, -2)$,
    \intervalgraph{inf}{open}{}{-2}
\end{answer}

---

6. $8 -3x \ge 5$

> \begin{answer}
$$
\begin{aligned}
8 - 3x &\ge 5\\
-3x &\ge -3\\
x &\le 1
\end{aligned}
$$
The interval is $(-\infty, 1]$,
    \intervalgraph{inf}{closed}{}{1}
\end{answer}

---

7. $5x - 3 \le 7 - 3x$

> \begin{answer}
$$
\begin{aligned}
5x - 3 &\le 7 - 3x\\
8x &\le 10\\
x &\le \frac{10}{8}
\end{aligned}
$$
The interval is $(-\infty, `r 5/4`]$,
    \intervalgraph{inf}{closed}{}{`r 5/4`}
\end{answer}

---

8. $3(2 - x) > 2(3 + x)$

> \begin{answer}
$$
\begin{aligned}
6 - 3x &> 6 + 2x\\
-3x &> 2x\\
0 &> 5x\\
x &< 0
\end{aligned}
$$
The interval is $(-\infty, 0)$,
    \intervalgraph{inf}{open}{}{0}
\end{answer}

---

9. $2x - \frac{1}{2} \ge 7x + \frac{7}{6}$

> \begin{answer}
$$
\begin{aligned}
2x - \frac{1}{2} &\ge 7x + \frac{7}{6}\\
2x - \frac{3}{6} &\ge 7x + \frac{7}{6}\\
-5x &\ge \frac{10}{6}\\
x &\le -\frac{1}{3}
\end{aligned}
$$
The interval is $\left(-\infty, -\frac{1}{3}\right]$,
    \intervalgraph{inf}{closed}{}{-1/3}
\end{answer}

---

10. $\frac{6 - x}{4} < \frac{3x - 4}{2}$

> \begin{answer}
$$
\begin{aligned}
\frac{6 - x}{4} &< \frac{3x - 4}{2}\\
6 - x &< 2(3x - 4)\\
6 - x &< 6x - 8\\
14 &< 7x\\
x &> 2
\end{aligned}
$$
The interval is $(2, \infty)$,
    \intervalgraph{open}{inf}{2}{}
\end{answer}

---

11. $\frac{4}{5}(x - 2) < \frac{1}{3}(x - 6)$

> \begin{answer}
$$
\begin{aligned}
\frac{12}{15}(x - 2) &< \frac{5}{15}(x - 6)\\
12x - 24 &< 5x - 30\\
12x &< 5x - 6\\
7x &< -6\\
x &< -\frac{6}{7}
\end{aligned}
$$
The interval is $\left(-\infty, -\frac{6}{7}\right)$,
    \intervalgraph{inf}{open}{}{-6/7}
\end{answer}

---

12. $-\frac{x + 5}{2} \le \frac{12 + 3x}{4}$

> \begin{answer}
$$
\begin{aligned}
-\frac{x + 5}{2} &\le \frac{12 + 3x}{4}\\
-2x - 10 &\le 12 + 3x\\
-5x &\le 22\\
x &\le -\frac{22}{5}
\end{aligned}
$$
The interval is $\left(-\infty, -22/5\right]$,
    \intervalgraph{inf}{closed}{}{-22/5}
\end{answer}

## Absolute Value

Solve the equations in Exercises 13--18.

---

13. $|y| = 3$

> \begin{answer}
$$
\begin{aligned}
|y| &= 3\\
y &= \pm3
\end{aligned}
$$
\end{answer}

---

14. $|y - 3| = 7$

> \begin{answer}
$$
\begin{aligned}
|y - 3| &= 7\\
|y| + |-3| &= 7\\
|y| + 3 &= 7\\
|y| &= 4\\
y &= \pm4
\end{aligned}
$$
\end{answer}

---

15. $|2t + 5| = 4$

> \begin{answer}
$$
\begin{aligned}
|2t + 5| &= 4\\
2t + 5 = 4 &\quad 2t + 5 = -4\\
2t = -1 &\quad 2t = -9\\
t = -\frac{1}{2} &\quad t = -\frac{9}{2}
\end{aligned}
$$
\end{answer}

---

16. $|1 - t| = 1$

> \begin{answer}
$$
\begin{aligned}
|1 - t| &= 1\\
1 - t = 1 &\quad 1 - t = -1\\
t = 0 &\quad t = 2
\end{aligned}
$$
\end{answer}

---

17. $|8 - 3s| = \frac{9}{2}$

> \begin{answer}
$$
\begin{aligned}
|8 - 3s| &= \frac{9}{2}\\
8 - 3s = \frac{9}{2} &\quad 8 - 3s = -\frac{9}{2}\\
3s = \frac{7}{2} &\quad 3s = \frac{25}{2}\\
s = \frac{7}{6} &\quad s = \frac{25}{6}
\end{aligned}
$$
\end{answer}

---

18. $\left|\frac{s}{2} - 1\right| = 1$

> \begin{answer}
$$
\begin{aligned}
\left|\frac{s}{2} - 1\right| &= 1\\
\frac{s}{2} - 1 = 1 &\quad \frac{s}{2} - 1 = -1\\
s = 4 &\quad s = 0
\end{aligned}
$$
\end{answer}

---

Solve the inequalities in Exercises 19--34, expressing the solution sets as
intervals or unions of intervals. Also, graph each solution set on the real
line.

19. $|x| < 2$

> \begin{answer}
$$
\begin{aligned}
|x| &< 2\\
-2 &< x < 2\\
\end{aligned}
$$
The interval $(-2, 2)$ \intervalgraph{open}{open}{-2}{2}
\end{answer}

---

20. $|x| \le 2$

> \begin{answer}
$$
\begin{aligned}
|x| &\le 2\\
-2 &\le x \le 2\\
\end{aligned}
$$
The interval $[-2, 2]$ \intervalgraph{closed}{closed}{-2}{2}
\end{answer}

---

21. $|t - 1| \le 3$

> \begin{answer}
$$
\begin{aligned}
|t - 1| &\le 3\\
-3 &\le t - 1 \le 3\\
-2 &\le t \le 4
\end{aligned}
$$
The interval $[-2, 4]$ \intervalgraph{closed}{closed}{-2}{4}
\end{answer}

---

22. $|t + 2| < 1$

> \begin{answer}
$$
\begin{aligned}
|t + 2| &< 1\\
-1 &< t + 2 < 1\\
-3 &< t < -1
\end{aligned}
$$
The interval $(-3, -1)$ \intervalgraph{open}{open}{-3}{-1}
\end{answer}

---

23. $|3y - 7| < 4$

> \begin{answer}
$$
\begin{aligned}
|3y - 7| &< 4\\
-4 &< 3y - 7 < 4\\
3 &< 3y < 11\\
1 &< y < \frac{11}{3}
\end{aligned}
$$
The interval $\left(1, \frac{11}{3}\right)$ \intervalgraph{open}{open}{1}{11/3}
\end{answer}

---

24. $|2y + 5| < 1$

> \begin{answer}
$$
\begin{aligned}
|2y + 5| &< 1\\
-1 &< 2y + 5 < 1\\
-6 &< 2y < -4\\
-3 &< y < -2\\
\end{aligned}
$$
The interval $(-3, -2)$ \intervalgraph{open}{open}{-3}{-2}
\end{answer}

---

25. $\left|\frac{z}{5} - 1\right| \le 1$

> \begin{answer}
$$
\begin{aligned}
\left|\frac{z}{5} - 1\right| &\le 1\\
-1 &\le \frac{z}{5} -1 \le 1\\
0 &\le \frac{z}{5} \le 2\\
0 &\le z \le 10
\end{aligned}
$$
The interval $[0, 10]$ \intervalgraph{closed}{closed}{0}{10}
\end{answer}

---

26. $\left|\frac{3}{2}z - 1\right| \le 2$

> \begin{answer}
$$
\begin{aligned}
\left|\frac{3}{2}z - 1\right| &\le 2\\
-2 &\le \frac{3}{2}z - 1 \le 2\\
-1 &\le \frac{3}{2}z \le 3\\
-2 &\le 3z \le 6\\
-\frac{2}{3} &\le z \le 2
\end{aligned}
$$
The interval $\left[-\frac{2}{3}, 2\right]$
    \intervalgraph{closed}{closed}{-2/3}{2}
\end{answer}

---

27. $\left|3 - \frac{1}{x}\right| < \frac{1}{2}$

> \begin{answer}
$$
\begin{aligned}
\left|3 - \frac{1}{x}\right| &< \frac{1}{2}\\
-\frac{1}{2} &< 3 - \frac{1}{x} < \frac{1}{2}\\
-\frac{7}{2} &< -\frac{1}{x} < -\frac{5}{2}\\
\frac{5}{2} &< x < \frac{7}{2}
\end{aligned}
$$
The interval $(5/2, 7/2)$ \intervalgraph{open}{open}{5/2}{7/2}
\end{answer}

---

28. $\left|\frac{2}{x} - 4\right| < 3$

> \begin{answer}
$$
\begin{aligned}
\left|\frac{2}{x} - 4\right| &< 3\\
-3 &< \frac{2}{x} - 4 < 3\\
1 &< \frac{2}{x} < 7\\
\frac{1}{2} &< \frac{1}{x} < \frac{7}{2}\\
\frac{2}{7} &< x < 2
\end{aligned}
$$
The interval $\left(\frac{2}{7}, 2\right)$
    \intervalgraph{open}{open}{2/7}{2}
\end{answer}

---

29. $|2s| \ge 4$

> \begin{answer}
$$
\begin{aligned}
|2s| &\ge 4\\
2s \le -4 &\quad 2s \ge 4\\
s \le -2 &\quad s \ge 2
\end{aligned}
$$
The intervals $(-\infty, -2]\cup[2, -\infty)$
    \intervalgraph{inf}{closed}{}{-2} \intervalgraph{closed}{inf}{2}{}
\end{answer}

---

30. $|s + 3| \ge \frac{1}{2}$

> \begin{answer}
$$
\begin{aligned}
|s + 3| &\ge \frac{1}{2}\\
s + 3 \le -\frac{1}{2} &\quad s + 3 \ge \frac{1}{2}\\
s \le -\frac{7}{2} &\quad s \ge -\frac{5}{2}
\end{aligned}
$$
The intervals $\left(-\infty, -\frac{7}{2}\right]
    \cup\left[-\frac{5}{2}, \infty\right)$
    \intervalgraph{inf}{closed}{}{-5/2} \intervalgraph{closed}{inf}{-7/2}{}
\end{answer}

---

31. $|1 - x| > 1$

> \begin{answer}
$$
\begin{aligned}
|1 - x| &> 1\\
1 - x < -1 &\quad 1 - x > 1\\
x > 2 &\quad x < 0
\end{aligned}
$$
The intervals $(-\infty, 0)\cup(2, \infty)$
    \intervalgraph{inf}{open}{}{0} \intervalgraph{open}{inf}{2}{}
\end{answer}

---

32. $|2 - 3x| > 5$

> \begin{answer}
$$
\begin{aligned}
|2 - 3x| &> 5\\
2 - 3x < -5 &\quad 2 - 3x > 5\\
-3x < -7 &\quad -3x > 3\\
x > \frac{7}{3} &\quad x < 1
\end{aligned}
$$
The intervals $(-\infty, 1)\cup(7/3, \infty)$
    \intervalgraph{inf}{open}{}{7/3} \intervalgraph{open}{inf}{1}{}
\end{answer}

---

33. $\left|\frac{r + 1}{2}\right| \ge 1$

> \begin{answer}
$$
\begin{aligned}
\left|\frac{r + 1}{2}\right| \ge 1\\
\frac{r + 1}{2} \le -1 &\quad \frac{r + 1}{2} \ge 1\\
r + 1 \le -2 &\quad r + 1 \ge 2\\
r \le -3 &\quad r \ge 1
\end{aligned}
$$
The intervals $(-\infty, -3]\cup[1, \infty)$
    \intervalgraph{inf}{closed}{}{-3} \intervalgraph{closed}{inf}{1}{}
\end{answer}

---

34. $\left|\frac{3r}{5} - 1\right| > \frac{2}{5}$

> \begin{answer}
$$
\begin{aligned}
\left|\frac{3r}{5} - 1\right| &> \frac{2}{5}\\
\frac{3r}{5} - 1 < -\frac{2}{5} &\quad \frac{3r}{5} - 1 > \frac{2}{5}\\
\frac{3r}{5} < \frac{3}{5} &\quad \frac{3r}{5} > \frac{7}{5}\\
3r < 3 &\quad 3r > 7\\
r < 1 &\quad r > \frac{7}{3}
\end{aligned}
$$
The intervals $(-\infty, 1)\cup(7/3, \infty)$
    \intervalgraph{inf}{open}{}{1} \intervalgraph{open}{inf}{7/3}{}
\end{answer}

## Quadratic Inequalities

Solve the inequalities in Exercises 35--42. Express the solution sets as
intervals or unions of intervals and graph them. Use the result $\sqrt{a^2} =
|a|$ as appropriate.

35. $x^2 < 2$

> \begin{answer}
$$
\begin{aligned}
x^2 &< 2\\
\sqrt{x^2} &< \sqrt{2}\\
|x| &< \sqrt{2}\\
-\sqrt{2} &< x < \sqrt{2}
\end{aligned}
$$
The intervals $(-\sqrt{2}, \sqrt{2})$
    \intervalgraph{open}{open}{-\sqrt{2}}{\sqrt{2}}
\end{answer}

---

36. $4 \le x^2$

> \begin{answer}
$$
\begin{aligned}
4 &\le x^2\\
\sqrt{4} &\le \sqrt{x^2}\\
2 &\le |x|\\
x \le -2 &\quad x \ge 2
\end{aligned}
$$
The intervals $(-\infty, -2]\cup[2, \infty)$
    \intervalgraph{inf}{closed}{}{-2} \intervalgraph{closed}{inf}{2}{}
\end{answer}

---

37. $4 < x^2 < 9$

> \begin{answer}
$$
\begin{aligned}
4 &< x^2 < 9\\
\sqrt{4} &< \sqrt{x^2} < \sqrt{9}\\
2 &< |x| < 3\\\\
|x| > 2 &\Rightarrow x < -2 \lor x > 2\\
|x| < 3 &\Rightarrow -3 < x < 3\\
x < -2 \land -3 < x < 3 &\Leftrightarrow -3 < x < -2\\
x > 2 \land -3 < x < 3 &\Leftrightarrow 2 < x < 3
\end{aligned}
$$
The intervals $(-3, -2)\cup(2, 3)$
  \intervalgraph{open}{open}{-3}{-2} \intervalgraph{open}{open}{2}{3}
\end{answer}

---

38. $\frac{1}{9} < x^2 < \frac{1}{4}$

> \begin{answer}
$$
\begin{aligned}
\frac{1}{9} < &x^2 < \frac{1}{4}\\
\sqrt{\frac{1}{9}} < &\sqrt{x^2} < \sqrt{\frac{1}{4}}\\
\frac{1}{3} < |x| &\land |x| < \frac{1}{2}\\\\
\frac{1}{3} < |x| &\Leftrightarrow x < -\frac{1}{3} \lor x > \frac{1}{3}\\
|x| < \frac{1}{2} &\Leftrightarrow -\frac{1}{2} < x < \frac{1}{2}\\
x < -\frac{1}{3} \land -\frac{1}{2} < x < \frac{1}{2} &\Leftrightarrow
    -\frac{1}{2} < x < -\frac{1}{3}\\
x > \frac{1}{3} \land -\frac{1}{2} < x < \frac{1}{2} &\Leftrightarrow
    \frac{1}{3} < x < \frac{1}{2}
\end{aligned}
$$
The intervals $(-\frac{1}{2}, -\frac{1}{3})\cup(\frac{1}{3}, \frac{1}{2})$
    \intervalgraph{open}{open}{-1/2}{-1/3} \intervalgraph{open}{open}{1/3}{1/2}
\end{answer}

---

39. $(x - 1)^2 < 4$

> \begin{answer}
$$
\begin{aligned}
(x - 1)^2 &< 4\\
\sqrt{(x - 1)^2} &< \sqrt{4}\\
|x - 1| &< 2\\
-2 &< x - 1 < 2\\
-1 &< x < 3
\end{aligned}
$$
The interval $(-1, 3)$
    \intervalgraph{open}{open}{-1}{3}
\end{answer}

---

40. $(x + 3)^2 < 2$

> \begin{answer}
$$
\begin{aligned}
(x + 3)^2 &< 2\\
\sqrt{(x + 3)^2} &< \sqrt{2}\\
|x + 3| &< \sqrt{2}\\
-\sqrt{2} &< x + 3 < \sqrt{2}\\
-\sqrt{2} - 3 &< x < \sqrt{2} - 3
\end{aligned}
$$
The interval $(-\sqrt{2}-3, \sqrt{2}-3)$
    \intervalgraph{open}{open}{-\sqrt{2}-3}{\sqrt{2}-3}
\end{answer}

41. $x^2 - x < 0$

> \begin{answer}
This problem can be solved via factoring or completing the square:\newline
\textbf{Method I (Factoring)}
$$
\begin{aligned}
x^2 - x &< 0\\
x(x - 1) &< 0\\
x < 0 \land x - 1 > 0 &\lor x > 0 \land x - 1 < 0\\
x < 0 \land x > 1 &\lor x > 0 \land x < 1\\
\text{only the second case is logical}&\\
0 < x < 1
\end{aligned}
$$
\textbf{Method II (Complete the square)}
$$
\begin{aligned}
x^2 - x &< 0\\
\left(x - \frac{1}{2}\right)^2 - \frac{1}{4} &< 0\\
\left(x - \frac{1}{2}\right)^2 &< \frac{1}{4}\\
\sqrt{\left(x - \frac{1}{2}\right)^2} &< \sqrt{\frac{1}{4}}\\
\left|x - \frac{1}{2}\right| &< \frac{1}{2}\\
-\frac{1}{2} < x - \frac{1}{2} < \frac{1}{2}\\
0 < x < 1
\end{aligned}
$$
The interval $(0, 1)$
    \intervalgraph{open}{open}{0}{1}
\end{answer}

42. $x^2 - x - 2 \ge 0$

> \begin{answer}
$$
\begin{aligned}
x^2 - x - 2 &\ge 0\\
\left(x - \frac{1}{2}\right)^2 - 2 - \frac{1}{4} &\ge 0\\
\left(x - \frac{1}{2}\right)^2 - \frac{9}{4} &\ge 0\\
\left(x - \frac{1}{2}\right)^2 &\ge \frac{9}{4}\\
\sqrt{\left(x - \frac{1}{2}\right)^2} &\ge \sqrt{\frac{9}{4}}\\
\left|x - \frac{1}{2}\right| &\ge \frac{3}{2}\\
x - \frac{1}{2} \ge \frac{3}{2} &\lor x - \frac{1}{2} \le -\frac{3}{2}\\
x \ge 2 &\lor x \le -1
\end{aligned}
$$
The intervals $(-\infty, -1]\cup[2, \infty)$
    \intervalgraph{inf}{closed}{}{-1} \intervalgraph{closed}{inf}{2}{}
\end{answer}

---

## Theory and Examples

43. Do not fall into the trap $|-a| = a$. For what real numbers $a$ is this
    equation true? For what real numbers is it false?

> \begin{answer}
This will hold true for all positive real numbers or zero, and is false for all
negative real numbers.
\end{answer}

---

44. Solve the equation $|x - 1| = 1 - x$.

> \begin{answer}
$$
\begin{aligned}
|x - 1| &= 1 - x\\
|x - 1| &= (1 - x)\\
&\Rightarrow x - 1 \le 0 (\text{since } |A| = -A \Leftrightarrow A \le 0)\\
&\Rightarrow x \le 1
\end{aligned}
$$
\end{answer}

---

45. __A proof of the triangle inequality.__ Give the reason justifying each of
    the numbered steps in the following proof of the triangle inequality.

$$
\begin{aligned}
|a + b|^2 &= (a + b)^2\\
&= a^2 + 2ab + b^2\\
&\le a^2 + 2|a||b| + b^2\\
&\le |a|^2 + 2|a||b| + |b|^2\\
&= (|a| + |b|)^2\\
|a + b| &\le |a| + |b|
\end{aligned}
$$

> \begin{answer}
Step-by-step breakdown.
\begin{enumerate}
\item $\sqrt{x^2} = |x| \Rightarrow \sqrt{x^2}^2 = |x|^2 \Rightarrow x^2 =
    |x|^2$.
\item $= a^2 + 2ab + b^2$. Binomial expansion of $(a + b)^2$.
\item $\le a^2 + 2|a||b| + b^2$. Because $ab \le |a||b|$. If $ab$ have opposite
    signs, the value is negative. When $ab$ have the same sign or one is zero,
    then the two sides are equal.
\item $\le |a|^2 + 2|a||b| + |b|^2$. Because $\sqrt{x^2} = |x| \Rightarrow
    \sqrt{x^2}^2 = |x|^2 \Rightarrow x^2 = |x|^2$.
\item $= (|a| + |b|)^2$. Factor $|a|^2 + 2|a||b| + |b|^2$.
\item $|a + b| \le |a| + |b|$. Take the square root of both sides, which
preserves the inequality since both sides are non-negative, and use $\sqrt{x^2}
= |x|$.
\end{enumerate}
\end{answer}

---

46. Prove that $|ab| = |a||b|$ for any numbers $a$ and $b$.

> \begin{answer}
$$
\begin{aligned}
|ab| = \sqrt{(ab)^2} = \sqrt{a^2 b^2} = \sqrt{a^2} \sqrt{b^2} = |a||b|
\end{aligned}
$$
\end{answer}

---

47. If $|x| \le 3$ and $x > -\frac{1}{2}$, what can you say about $x$?

> \begin{answer}
There are two restrictions on the value of $x$.
$$
\begin{aligned}
-3 \le x \le 3 \land x > -\frac{1}{2} \Rightarrow -\frac{1}{2} < x \le 3.
\end{aligned}
$$
\end{answer}

---

48. Graph the inequality $|x| + |y| \le 1$.

> \begin{answer}
Here is the logic.
\begin{enumerate}[label=Step \arabic*:]
\item Split by sign, since $|x|$ and $|y|$ each depend on sign. There are four
      regions (quadrants) to consider, based on the signs of $x$ and $y$:
    \begin{itemize}
    \item \textbf{Quadrant I} ($x \ge 0, y \ge 0$): $|x| = x,|y| = y$, so the
        inequality becomes $x + y \le 1$.
    \item \textbf{Quadrant II} ($x \le 0, y \ge 0$): $|x| = -x,|y| = y$, so it
        becomes $-x + y \le 1$.
    \item \textbf{Quadrant III} ($x \le 0, y \le 0$): $|x| = -x,|y| = -y$, so it
        becomes $-x-y \le 1$.
    \item \textbf{Quadrant IV} ($x \ge 0, y \le 0$): $|x| = x,|y| = -y$, so it
        becomes $x - y \le 1$.
    \end{itemize}
\item Each linear inequality is a half-plane. In each quadrant, $|x| + |y| \le
      1$ reduces to an ordinary linear inequality, so within that quadrant the
      boundary $|x| + |y| = 1$ is a straight line segment:
    \begin{itemize}
    \item Quadrant I: $x + y = 1$--a line through (1,0) and (0,1).
    \item Quadrant II: $-x + y = 1$--a line through (0,1) and (-1,0).
    \item Quadrant III: $-x-y=1$--a line through (-1,0) and (0,-1).
    \item Quadrant IV: $x - y=1$--a line through (0, -1) and (1,0).
    \end{itemize}
\item Stich the four segments together. Each pice is a straight edge connecting
      two adjacent axis intercepts: $(1,0) \rightarrow (0,1) \rightarrow (-1,0)
      \rightarrow (0,-1) \rightarrow (1,0)$. Four straight edges meeting at four
      vertices is exactly a \textbf{diamond} (a square rotated $45^{\circ}$),
      which is why the boundary in the code is the `polygon()` with those four
      vertices.
\item Why "$\le$" fills the interior. In each quadrant, the inequality is
      (linear expression) $\le 1$, which describes the half-plane
      \textbf{containing the origin} (since plugging in $x=y=0$ gives $0 \le 1$,
      true). So the solution set is everything the diamond, consistent with
      filling the polygon rather than just drawing its outline.
\end{enumerate}
\end{answer}

```{r ex48, echo=F, fig.height=4, fig.width=4}
x <- seq(-1.5, 1.5, length.out = 400)
y <- seq(-1.5, 1.5, length.out = 400)

plot(0, 0, type = "n", xlim = c(-1.5, 1.5), ylim = c(-1.5, 1.5),
     xlab = "x", ylab = "y", asp = 1)
abline(h = 0, v = 0, col = "gray")

# The boundary |x| + |y| = 1 is a diamond (rotated square)
# with vertices at (1,0), (0,1), (-1, 0), (0,-1)
vertices_x <- c(1, 0, -1, 0, 1)
vertices_y <- c(0, 1, 0, -1, 0)

polygon(vertices_x, vertices_y, col = "lightblue", border = "blue", lwd = 2)
```

---

49. GRAPHER

> a) Graph the functions $f(x) = x/2$ and $g(x) = 1 + (4/x)$ together to
  identify the values of $x$ for which
$$
\begin{aligned}
\frac{x}{2} > 1 + \frac{4}{x}.
\end{aligned}
$$

> b) Confirm your findings in (a) algebraically.

```{r ex49, echo=F, fig.height=4, fig.width=4}
x.1 <- seq(-5, -0.1, length.out = 500)
x.2 <- seq(0.1, 5, length.out = 500)
f.1 <- x.1 / 2; g.1 <- 1 + 4 / x.1
f.2 <- x.2 / 2; g.2 <- 1 + 4 / x.2
plot(0, 0, type = "n", xlim = c(-5, 5), ylim = c(-5, 5),
    xlab = "x", ylab = "y", asp = 1)
abline(h = 0, v = 0, col = "gray")
lines(x.1, f.1);lines(x.1, g.1, col = "red")
lines(x.2, f.2);lines(x.2, g.2, col = "red")
```

> \begin{answer}
$$
\begin{aligned}
\frac{x}{2} &> 1 + \frac{4}{x}\\
x &> 2 + \frac{8}{x}\\
\end{aligned}
$$
\begin{enumerate}
\item[Case $x>0$:] multiplying by $2x$ (positive) preserves the inequality:
$$
\begin{aligned}
x^2 > 2x + 8 \Rightarrow (x - 4)(x + 2) > 0
    \Rightarrow x > 4 \lor x < -2
\end{aligned}
$$
\item[Case $x<0$:] multiplying by $2x$ (negative) reverses the inequality:
$$
\begin{aligned}
x^2 > 2x + 8 \Rightarrow (x - 4)(x + 2) < 0
    \Rightarrow -2 < x < 4
\end{aligned}
$$
\end{enumerate}
Combining both cases: $(-2, 0)\cup(4, \infty)$.
    \intervalgraph{open}{open}{-2}{0} \intervalgraph{open}{inf}{4}{}
\end{answer}

50. GRAPHER

> a) Graph the function $f(x) = \frac{3}{x-1}$ and $g(x) = \frac{2}{x + 1}$
  together to identify the values of $x$ for which

$$
\begin{aligned}
\frac{3}{x - 1} < \frac{2}{x + 1}
\end{aligned}
$$

```{r ex50, echo=F, fig.height=4, fig.width=4}
x.1 <- seq(-10, -1.1, length.out = 300)
x.2 <- seq(-0.9, 0.9, length.out = 300)
x.3 <- seq( 1.1,  10, length.out = 300)
f.1 <- 3 / (x.1 - 1);g.1 <- 2 / (x.1 + 1)
f.2 <- 3 / (x.2 - 1);g.2 <- 2 / (x.2 + 1)
f.3 <- 3 / (x.3 - 1);g.3 <- 2 / (x.3 + 1)
plot(0, 0, type = "n", xlim = c(-10, 10), ylim = c(-10, 10),
    xlab = "x", ylab = "y")
abline(h = 0, v = 0, col = "gray")
abline(v = c(-1, 1), col = "gray", lty = 2)
lines(x.1, f.1);lines(x.1, g.1, col = "red")
lines(x.2, f.2);lines(x.2, g.2, col = "red")
lines(x.3, f.3);lines(x.3, g.3, col = "red")
```

\begin{answer}
$$
\begin{aligned}
\frac{3}{x - 1} &< \frac{2}{x + 1}\\
\frac{3}{x - 1} - \frac{2}{x + 1} &< 0\\
\frac{3(x + 1) - 2(x - 1)}{(x - 1)(x + 1)} &< 0\\
\frac{x + 5}{x^2 - 1} &< 0
\end{aligned}
$$
We have four reagions:
\begin{itemize}
\item $x < -5$: numerator $-$, denominator $+$ (since $|x| > 1$) $\rightarrow$
    quotient $-$ $\rightarrow$ \textbf{satisfies} $< 0$.
\item $-5 < x < -1$: numerator $+$, denominator $+$ $\rightarrow$ quotient $+$
    $\rightarrow$ fails.
\item $-1 < x < 1$: numerator $+$, denominator $-$ (since $x^2 < 1$)
    $\rightarrow$ quotient $-$ $\rightarrow$ \textbf{satisfies} $< 0$.
\item $x > 1$: numerator $+$, denominator $+$ $\rightarrow$ quotient $+$
    $\rightarrow$ fails.
\end{itemize}
The solution set is $(-\infty, -5)\cup(-1, 1)$
\end{answer}
