---
output: pdf_document
title: "Chapter 8 Infinite Series"
toc: true
---

```{r setup, include=F}
require(tidyverse)
```

# 8.1 Limits of Sequences of Numbers

## Definitions and notation

> __Definition__<p>
> An __infinite sequence__ (or __sequence__) of numbers is a function whose domain
> is the set of integers greater than or equal to some integer $n_0$.

Usually $n_0$ is 1 and the domain of the sequence is the set of positive
integers. But sometimes we want to start sequences elsewhere. We take $n_0 = 0$
when we begin Newton's method. We might take $n_0 = 3$ if we were defining a
sequence of $n$-sided polygons.

Sequences are defined the way other functions are, some typical rules being

$$
\begin{aligned}
a(n) = \sqrt{n},\ \ \ & a(n) = (-1)^{n+1}\frac{1}{n},\ & a(n) = \frac{n - 1}{n}
\end{aligned}
$$

(Example 1 and Fig. 8.1).

To indicate that the domains are sets of integers, we use a letter like $n$ from
the middle of the alphabet for the independent variable, instead of the
$x, y, z$, and $t$ used widely in other contexts. The formulas in the defining
rules, however, like those above, are often valid for domains larger than the
set of positive integers. This can be an advantage as we will see.

The number $a(n)$ is the __$n$th term__ of the sequence, or the
__term with index $n$__. If $a(n) = (n - 1)/n$, we have:

$$
a(1) = 0,\ \ a(2) = \frac{1}{2},\ \ a(3) = \frac{2}{3},\ \ \ldots\ \
  a(n) = \frac{n - 1}{n}.
$$

When we use the subscript notation $a_n$ for $a(n)$, the sequence is written

$$
a_1 = 0,\ \ a_2 = \frac{1}{2},\ \ a_3 = \frac{2}{3},\ \ \ldots\ \
  a(n) = \frac{n - 1}{n}.
$$

To describe sequences, we often write the first few terms as well as a formula
for the $n$th term.

#### EXAMPLE 1

$$
\begin{aligned}
& \textbf{We write} & \textbf{For the sequence whose defining rule is} \\
& 1, \sqrt{2}, \sqrt{3}, \sqrt{4}, \ldots, \sqrt{n}, \ldots
  & a_n = \sqrt{n}\\\\
& 1, \frac{1}{2}, \frac{1}{3}, \ldots, \frac{1}{n}, \ldots
  & a_n  = \frac{1}{n}\\\\
& 1, -\frac{1}{2}, \frac{1}{3}, -\frac{1}{4}, \ldots,
    (-1)^{n+1} \frac{1}{n}, \ldots & a_n = (-1)^{n+1} \frac{1}{n}\\\\
& 0, \frac{1}{2}, \frac{2}{3}, \frac{3}{4}, \ldots, \frac{n - 1}{n}, \ldots
  & a_n = \frac{n - 1}{n}\\\\
& 0, -\frac{1}{2}, \frac{2}{3}, -\frac{3}{4}, \ldots,
    (-1)^{n+1}\left(\frac{n - 1}{n}\right), \ldots
  & a_n = (-1)^{n+1}\left(\frac{n - 1}{n}\right)\\\\
& 3, 3, 3, \ldots, 3, \ldots & a_n = 3
\end{aligned}
$$

#### Notation

We refer to the sequence whose $n$th term is $a_n$ with the notation $\{a_n\}$
("the sequence $a$ sub $n$"). The second sequence in Example 1 is $\{1/n\}$
("the sequence 1 over $n$"); the last sequence is $\{3\}$ ("the constant
sequence 3").

#### Fig. 8.1

The sequences of Example 1 are graphed here in two different ways; by plotting
the numbers $a_n$ on a horizontal axis and by plotiting the points ($n, a_n$) in
the coordinate plane.

```{r fig. 8.1 graph function, echo=F}
f8.1 <- new.env()
f8.1$graph <- function(f, n=1:5, exp=""){
    df <- tibble(
      x = n,
      y = f(x),
    )
    df$lab.1 <- paste0("a[", df$x, "]")
    df$lab.2 <- paste0("(", df$x, ", ", round(df$y, 2), ")")
    plot.1 <- ggplot(df, aes(x=y, y=0)) +
        geom_point() +
        geom_text(aes(label = lab.1), parse = TRUE, vjust = -0.5, size = 3) +
        theme_minimal() +
        ylab("") +
        xlab(exp) + 
        coord_equal(xlim = c(min(df$y)-1, max(df$y)+1), ylim = c(-.5, .5)) +
        theme(axis.text.y = element_blank(),
            axis.ticks.y = element_blank(),
            panel.grid.major.y = element_blank(),
            panel.grid.minor.y = element_blank())
    plot.2 <- ggplot(df, aes(x=x, y=y)) +
      geom_point() +
      geom_text(aes(label = lab.2), vjust = -0.5, size = 2) +
      coord_equal(xlim = c(0, length(n)+1), ylim = c(min(df$y)-1, (max(df$y)+1))) +
      ylab(expression(a[n])) +
      xlab("n") +
      theme_minimal()

   list(p1 = plot.1, p2 = plot.2)
}
```

The terms $a_n = \sqrt{n}$ eventually surpass every integer, so the sequence
$\{a_n\}$ diverges, \ldots

```{r graph 1, echo=F, fig.height=3, fig.width=3}
one <- f8.1$graph(function(x) sqrt(x), exp=expression(a[n] == sqrt(n)))
one$p1;one$p2
```

... but the terms $a_n = 1/n$ decrease steadily and get arbitrarily close to 0
as $n$ increases, so the sequence $\{a_n\}$ converges to 0.

```{r graph 2, echo=F, fig.height=3, fig.width=3}
two <- f8.1$graph(function(x) 1 / x, exp=expression(a[n] == frac(1, n)))
two$p1;two$p2
```

The terms $a_n = (-1)^{n+1}(1/n)$ alternate in sign but still converge to 0.

```{r graph 3, echo=F, fig.height=3, fig.width=3}
three <- f8.1$graph(function(x) (-1)^(x + 1) * 1 / x)
three$p1;three$p2
```

The terms $a_n = (n - 1)/n$ approach 1 steadily and get arbitrarily close as $n$
increases, so the sequence $\{a_n\}$ converges to 1.

```{r graph 4, echo=F, fig.height=3, fig.width=3}
four <- f8.1$graph(function(x) (x - 1) / x,
    exp=expression(a[n] == (-1)^(n+1) * frac(1, n)))
four$p1;four$p2
```

The terms $a_n = (-1)^{n + 1}[(n - 1)/n]$ alternate in sign. The positive terms
approach 1. But the negative terms approach -1 as $n$ increases, so the sequence
$\{a_n\}$ diverges.

```{r graph 5, echo=F, fig.height=3, fig.width=3}
five <- f8.1$graph(function(x) (-1)^(x+1)*((x - 1) / x))
five$p1;five$p2
```

The terms in the sequence of constants $a_n = 3$ have the same value regardless
of $n$, so the sequence $\{a_n\}$ converges to 3.

```{r graph 6, echo=F, fig.height=3, fig.width=3}
six <- f8.1$graph(function(x) 3)
six$p1;six$p2
```

## Convergence and Divergence

As Fig. 8.1 shows, the sequences of Example 1 do not behave the same way. The
sequences $\{1/n\}$, $\{(-1)^{n+1}(1/n)\}$, and $\{(n-1)/n]\}$ each seem to
approach a single limiting value as $n$ increases, and $\{3\}$ is at a limiting
value from the very first. On the other hand, terms of $\{(-1)^{n+1}(n-1)/n\}$
seem to accumulate near two different values, -1 and 1, while the terms of
$\{\sqrt{n}\}$ become increasingly large and do not accumulate anywhere.

To distinguish sequences that approach a unique limiting value $L$, as $n$
increases, from those that do not, we say that the former sequences _converge_,
according to the following definition.

> __Definition__<p>
> The sequence $\{a_n\}$ __converges__ to the number $L$ if to every positive
> number $\epsilon$ there corresponds an integer $N$ such that for all $n$,

$$
n > N\ \ \ \ \Rightarrow\ \ \ \ |a_n - L| < \epsilon
$$

> If no such number $L$ exists, we say that $\{a_n\}$ __diverges__.

> If $\{a_n\}$ converges to $L$, we write
> $\lim_{n \rightarrow \infty}{a_n} = > L$, or simply $a_n \rightarrow L$, and
> call $L$ the __limit__ of the sequence (Fig. 8.2).

#### Fig. 8.2

$a_n \rightarrow L$ if $y = L$ is a horizontal asymptote of the sequence of
points $\{(n, a_n)\}$. In this figure, all the $a_n$'s after $a_N$ lie within
$\epsilon$ of $L$.

```{r figure 8.2 line, echo=F, fig.height=3}
f8.2 <- new.env()
n <- 1:10
f8.2$df <- tibble(
    x = n,
    y = c(2, 1, 1.25, 1.75, 1.6, 2.1, 2.85, 3, 3.5, 3.3)
)
f8.2$points.labeled <- f8.2$df[c(6, 8),]
f8.2$lab <- paste0("a[", f8.2$df$x, "]")
f8.2$e <- .65
f8.2$L <- 3.4
f8.2$limits <- tibble(
    p = c(f8.2$L - f8.2$e, f8.2$L, f8.2$L + f8.2$e),
    v = c(0, 0, 0),
    l = c("(", "|", ")"),
    lt = c("1", "2", "1")
)
ggplot(f8.2$df, aes(x=y, y=0)) +
    geom_point() +
    geom_text(aes(label = f8.2$lab), parse = TRUE, vjust = -0.5, size = 3) +
    theme_minimal() +
    ylab("") +
    xlab("") + 
    coord_equal(xlim = c(min(f8.2$df$y)-1, max(f8.2$df$y)+1), ylim = c(-.5, .5)) +
    geom_hline(yintercept = 0, linewidth = .2) +
    geom_text(data = f8.2$limits, aes(x=p, y=v, label = l), size=10) +
    theme(axis.text.y = element_blank(),
        axis.ticks.y = element_blank(),
        panel.grid.major.y = element_blank(),
        panel.grid.minor.y = element_blank())
ggplot(f8.2$df, aes(x=x, y=y)) +
    geom_point(size = .8, color = "blue") +
    coord_equal(xlim = c(0, 12), ylim = c(0, 6)) +
    geom_hline(data=f8.2$limits, aes(yintercept = p, linetype = lt),
        color = "red", linewidth = .2) +
    scale_x_continuous(breaks = c(1, 2, 3, 6, 7, 8),
        labels = c(1, 2, 3, 'N', 7, 'n')) +
    geom_text(data = f8.2$points.labeled,
        aes(label = c("(N, a[N])", "(n, a[n])")),
        vjust = -1, size = 2)
```

#### EXAMPLE 2

Testing the definition

Show that

##### a.

$\lim_{n \rightarrow \infty}{\frac{1}{n}} = 0$

Let $\epsilon > 0$ be given. We must show that there exists an integer $N$ such
that for all $n$,

$$
n > N\ \ \ \ \Rightarrow\ \ \ \ \left|\frac{1}{n} - 0\right| < \epsilon.
$$

This implication will hold if $(1/n) < \epsilon$ or $n > 1/\epsilon$. If $N$ is
any integer greater than $1/\epsilon$, the implication will hold for all $n >
N$. This proves that $lim_{n \rightarrow \infty}{(1/n)} = 0$.

##### b.

$\lim_{n \rightarrow \infty}{k} = k$ (any constant $k$)

Let $\epsilon > 0$ be given. We must show that there exists an integer $N$
such that for all $n$,

$$
n > N\ \ \ \ \Rightarrow\ \ \ \ |k - k| < \epsilon.
$$

Since $k - k = 0$, we can use any positive integer for $N$ and the implication
will hold. This proves that $\lim_{n \rightarrow \infty}{k} = K$ for any
constant $k$.

#### EXAMPLE 3

Show that $\{(-1)^{n+1}[(n-1)/n]\}$ diverges.

Take a positive $\epsilon$ smaller than 1 so that the bands shown in Fig. 8.3
about the lines $y = 1$ and $y = -1$ do not overlap. Any $\epsilon < 1$ will do.
Convergence to 1 would require every point of the graph beyond a certain index
$N$ to lie inside the upper band, but this will never happen. As soon as a point
$(n, a_n)$ lies in the upper band, every alternate point starting with
$(n+1, a_{n+1})$ will lie in the lower band. Hence the sequence cannot converge
to 1. Likewise, it cannot converge to -1. On the other hand, because the terms
of the sequence get alternately closer to 1 and -1, they never accumulate near
any other value. Therefore, the sequence diverges.

The behavior of $\{(-1)^{n+1}[(n-1)/n]\}$ is qualitatively different from that
of $\{\sqrt{n}\}$, which diverges because it outgrows every real number $L$. To
describe the behavior of $\{\sqrt{n}\}$ we write

$$
\lim_{n \rightarrow \infty}{(\sqrt{n})} = \infty.
$$

In speaking of infinity as a limit of a sequence $\{a_n\}$, we do not mean that
the difference between $a_n$ and infinity becomes small as $n$ increases. We
mean that $a_n$ becomes numerically large as $n$ increases.

## Recursive Definitions

So far, we have calculated each $a_n$ directly from the value of $n$. But
sequences are often defined __recursively__ by giving

1. The value(s) of the initial term or terms, and

2. A rule, called a __recursion formula__, for calculating any later term from
   terms that precede it.

#### EXAMPLE 4

Sequences constructed recursively

a. The statements $a_1 = 1$ and $a_n = a_{n-1} + 1$ define the sequence
   $1, 2, 3, \ldots, n, \ldots$ of positive integers. With $a_1 = 1$, we have
   $a_2 = a_1 + 1 = 2$, $a_3 = a_2 + 1 = 3$, and so on.

b. The statements $a_1 = 1$ and $a_n = n \cdot a_{n-1}$ define the sequence
   $1, 2, 6, 24, \ldots, n!, \dots$ of factorials. With $a_1 = 1$, we have
   $a_2 = 2 \cdot a_1 = 2$, $a_3 = 3 \cdot a_2 = 6$, $a_4 = 4 \cdot a_3 = 24$,
   and so on.

c. The statements $a_1 = 1$, $a_2 = 1$, and $a_{a+1} = a_n + a_{n - 1}$ define
   the sequence $1, 1, 2, 3, 5, \ldots$ of __Fibonacci numbers__. With $a_1 = 1$
   and $a_2 = 1$, we have $a_3 = 1 + 1 = 2$, $a_4 = 2 + 1 = 3$,
   $a_5 = 3 + 2 = 5$, and so on.

d. As we can see by applying Newton's method, the statements $x_0 = 1$ and
   $x_{n+1} = x_n - [(\sin x_n - x_n^2) / (\cos x_n - 2x_n)]$ define a sequence
   that converges to a solution of the equation $\sin x - x^2 = 0$.

## Subsequences

If the terms of one sequence appear in another sequence in their given order, we
call the first sequence a subsequence of the second.
