---
title: "Chapter 1.1 Limits and Continuity"
output:
  pdf_document:
    toc: true
---

```{r; setup}
require(plotrix)
```

# Limits from graphs

## 1. graph 1

For the function $g(x)$ graphed here, find the following limits or explain
why they do not exist.

```{r problem 1; setup, fig.width=4, fig.height=3, echo=F}
plot(x <- seq(0, .96, by=.01), x, col="blue",
    bty="n", type='l',
    xlim=c(-.05, 4), ylim=c(-.05, 2),
    xlab="", ylab=""
)
lines(c(1, 2), c(0, 1), col="blue")
lines(c(2.06, 2.94), c(1, 0.06), col="blue")
lines(c(3.06, 4), c(0.06, 1), col="blue")
points(1, 1, pch=1, col="blue")
points(1, 0, pch=16, col="blue")
points(2, 1, pch=16, col="blue")
points(3, 1, pch=16, col="blue")
points(3, 0, pch=1, col="blue")
lines(c(0, 3), c(1, 1), col="red", lty=2)
```

---

### A. $\lim_{x \rightarrow 1}{g(x)}$

The limit does not exist. As $x$ approaches 1 from the right, $g(x)$ approaches
0, As $x$ approaches 1 from the left, $g(x)$ approaches 1. There is no single
limit $L$ that all values of $g(x)$ approaches, as x approahces 1.

### B. $\lim_{x \rightarrow 2}{g(x)}$

L = 2. The limit is defined on both sides.

### C. $\lim_{x \rightarrow 3}{g(x)}$

L = 0. The limit is defined for both sides. Even though the function defines
$g(x) = 1$ at $x = 3$, The limit of $g(x)$ approaches 0 as $x$ approaches 3.

---

## 2. Graph 2

For the function $f(x)$ graphed here, find the following limits or explain why
they do not exist.

```{r problem 2; setup, fig.width=4, fig.height=3, echo=F}
plot(x <- seq(-3, -2.06, by=.01), x + 2, col="blue",
     bty="n", type="l",
     xlim=c(-3,2), ylim=c(-2, 2),
     xlab="", ylab="")
lines(c(-1.94, -1.06), c(-.06, -.94), col="blue")
lines(c(-.94, -.06), c(-1, -1), col="blue")
lines(c(.06, 2), c(1, 1), col="blue")
points(0, -1, pch=1, col="blue")
points(0, 1, pch=16, col="blue")
points(-1, -1, pch=16, col="blue")
points(-2, 0, pch=1, col="blue")
```

---

### A. $\lim_{t \rightarrow -2}{f(t)}$

L=0. The limit is defined on both sides.

### B. $\lim_{t \rightarrow -1}{f(t)}$

L=-1. The limit is defined on both sides.

### C. $\lim_{t \rightarrow 0}{f(t)}$

The limit does not exist because the limit is different for each side.

---

## 3. Graph 3

Which of the following statements about the function $y = f(x)$ graphed here are
true, and which are false?

```{r problem 3; setup, fig.width=4, fig.height=3, echo=F}
plot(x <- seq(-1, -.06, by=.01), x, col="blue",
     bty="n", type="l",
     xlim=c(-1.5,1.5), ylim=c(-1.5, 1.5),
     xlab="", ylab="")
abline(h=0, col="gray")
abline(v=0, col="gray")
lines(c(.06, .94), c(-.06, -.96), col="blue")
lines(c(1, 2), c(0, 1), col="blue")
points(-1, -1, pch=16, col="blue")
points(0, 0, pch=1, col="blue")
points(0, 1, pch=16, col="blue")
points(1, -1, pch=1, col="blue")
points(1, 0, pch=16, col="blue")
points(2, 1, pch=16, col="blue")
```

---

### A. $\lim_{x \rightarrow 0}{f(x)}$ exists

True, This limit exists, because approaching the value $x = 0$ from both sides
gives the same result $L = 0$.

### B. $\lim_{x \rightarrow 0}{f(x) = 0}$

True

### C. $\lim_{x \rightarrow 0}{f(x) = 1}$

False, this is the true value, but the limit is $L = 0$.

### D. $\lim_{x \rightarrow 1}{f(x) = 1}$

False

### E. $\lim_{x \rightarrow 1}{f(x) = 0}$

False, this limit is undefined becuase approaching $x$ from either side of 1
gives different results.

### F. $\lim_{x \rightarrow x_0}{f(x)}$ exists at every point $x_0$ in (-1, 1)

True. Disregarding $x=1$, which is not part of the interval.

---

## 4. Graph 4

Which of the following statements about the function $y = f(x)$ graphed here are
true, and which are false?

```{r problem 4; setup, fig.width=4, fig.height=3, echo=F}
#xlim <- c(-1.5, 3.5)
#ylim <- c(-2.5, 1.5)
#x <- seq(-1, -.06, by=.01)
#plot(x, x, col="blue",
#     bty="n", type="l",
#     xlim=xlim, ylim=ylim,
#     xlab="", ylab="")
#abline(h=0, col="gray")
#abline(v=0, col="gray")
plot(c(-10,10), c(-10,10), type="n",
     bty="n", xlab="", ylab="")
plotrix::draw.ellipse(c(0, 0), c(0,0), 5.2, 8.8, lwd=1,
    segment=c(0, -45), nv=c(3,4), arc.only=T)
#lines(c(1, 2), c(0, 1), col="blue")
#points(-1, -1, pch=16, col="blue")
#points(0, 0, pch=1, col="blue")
#points(0, 1, pch=16, col="blue")
#points(1, -1, pch=1, col="blue")
#points(1, 0, pch=16, col="blue")
#points(2, 1, pch=16, col="blue")
```

---

### A. $\lim_{x \rightarrow 2}{f(x)}$ does not exist.

False. The limit exists as $L = 1$ even though the function defines f(2)=0.

### B. $\lim_{x \rightarrow 2}{f(x)} = 2$

False. The limit $L = 1$ as $x \rightarrow 2$.

### C. $\lim_{x \rightarrow 1}{f(x)}$ does not exist.

True. As you approach $x \rightarrow 1$ from either side, the limit is
different.

### D. $\lim_{x \rightarrow x_0}{f(x)}$ exists at every point $x_0$ in (-1, 1)

True.

### E. $\lim_{x \rightarrow x_0}{f(x)}$ exists at every point $x_0$ in (1, 3)

True.

---

# Existence of limits

In Exercises 5 and 6, explain why the limits do not exist.

## 5. $\lim_{x \rightarrow 0}{\frac{x}{|x|}}$

```{r problem 5; setup, echo=F}
plot(x <- seq(-1, 1, by=.01), x / abs(x), type='l', col="blue")
```

As you approach $x \rightarrow 0$ from either side you get different results.

---

## 6. $\lim_{x \rightarrow 1}{\frac{1}{x - 1}}$

```{r problem 6; setup, echo=F}
plot(x <- seq(0, 2, by=.01), 1 / (x - 1), type='l', col="blue")
```

The function $f(x)$ is not defined at $x=1$. When approaching from the left you
go to $-\infty$ and when you approach from the right you go to
$\infty$.

---

## 7. The existence of limits

Suppose that a function $f(x) is defined for all real values of $x$ except
$x = x_0$. Can anything be said about the existence of
$\lim_{x \rightarrow x_0}{f(x)}$? Give reasons for your answer.

---

The existence of $\lim_{x \rightarrow x_0}{f(x)}$ does not depend on whether
$f(x_0)$ is defined.

The limit exists if and only if the values of $f(x)$ get arbitrarily close to
some number $L$ as $x$ gets arbitrarily close to $x_0$, from both
sides---regarless of whether $f(x_0)$ is defined.

---

## 8. The existence of limits

Suppose that a function $f(x)$ is defined for all $x$ in [-1, 1]. Can anything
be said about the existence of $\lim_{x \rightarrow 0}{f(x)}$? Give reasons for
your answer.

---

As stated above, the existence of a limit at a particular point does not depend on whether the
function is defined at that point. The limit's existence depends on getting
arbitrarily close to some number $L$, as you get close to $x_0$.

---

## 9. The existence of limits

If $\lim_{x \rightarrow 1}{f(x)} = 5$, must $f$ be defined at $x=1$? If it is,
must $f(1) = 5$? Can we conclude _anything_ about the values of $f$ at $x = 1$?
Explain.

Answer is the same as above. Additionally, $f(1)$ does not have to equal $5$ in
order for the limit $L = 5$. The reason is the same as above.

---

## 10. The existence of limits

If $f(1) = 5$, must $\lim_{x \rightarrow 1}{f(x)}$ exist? If it does, then must
$\lim_{x \rightarrow 1}{f(x)} = 5$? Can we conclude _anything_ about
$\lim_{x \rightarrow 1}{f(x)}$? Explain.

---

This is the same answer as above.

---

# Calculator/Grapher exercises---Estimating limits

## 11. Let $f(x) = (x^2 - 9)/(x + 3)$

a) __CALCULATOR__ Make a table of the values of $f$ at the points
   $x = -3.1, -3.01, -3.001$, and so on as far as your calculator can go. Then
   estimate $\lim_{x \rightarrow -3}{f(x)}$. What estimate do you arrive at if
   you evaluate $f$ at $x = -2.9, -2.99, -2.999, \ldots$ instead?

b) __GRAPHER__ Support your conclusions in (a) by graphing $f$ near $x_0 = -3$.

c) Find $\lim_{x \rightarrow -3}{f(x)}$ algebraically.

---

### A. Table

```{r problem 11; table, echo=F}
x.rt <- c(-3.1, -3.01, -3.001, -3.0001,
          -3.00001, -3.000001, -3.0000001,
          -3.00000001, -3.000000001, -3.0000000001)
x.lt <- c(-2.9, -2.99, -2.999, -2.9999,
          -2.99999, -2.999999, -2.9999999,
          -2.99999999, -2.999999999,
          -2.9999999999)
f <- function(x) (x^2 - 9)/(x + 3)
f.x.rt <- f(x.rt)
f.x.lt <- f(x.lt)
knitr::kable(data.frame(
    X = x.rt,
    f = f.x.rt), caption="From the right")
knitr::kable(data.frame(
    X = x.lt,
    f = f.x.lt), caption="From the left")
```

### B. Graph

```{r problem 11; graph, echo=F}
x <- seq(-3.1, -2.9, by=.001)
plot(x, f(x), pch=19, cex=.02)
points(-3, -6, cex=.8, col="blue")
```

There is a hole at $f(-3) = -6$.

---

## 12. Let $g(x) = (x^2 - 2) / (x - sqrt(2))$

a) __CALCULATOR__ Make a table of the values of $g$ at the points
   $x = 1.4, 1.41, 1.414, \ldots$, and so on through successive decimal
   approximations of $\sqrt{2}$. Estimate $\lim_{x \rightarrow \sqrt{2}}{g(x)}$.

b) __GRAPHER__ Support your conclusion in (a) by graphing $g$ near
   $x_0 = \sqrt{2}$.

c) Find $\lim_{x \rightarrow \sqrt{2}}{g(x)}$ algebraically.
