---
title: 1.3 Target Values and Formal Definition of Limits
output:
  pdf_document:
    toc: true
    includes:
      in_header: "../preamble.tex"
---

```{r setup, message=F, echo=F}
library(ggplot2)
library(patchwork)
library(latex2exp)

plot_interval <- function(a, b, x0, label_a = "a", type = "open",
                          label_b = "b", label_x0 = TeX("$x_0$")) {

    stopifnot(a < x0, x0 < b)

    pch_a <- pch_b <- switch(type,
        "open"        = 1,
        "closed"      = 19,
        "open-closed" = c(1, 19),
        "closed-open" = c(19, 1)
    )

    # for the mixed cases, split into two values
    if (type == "open-closed") { pch_a <- 1;  pch_b <- 19 }
    if (type == "closed-open") { pch_a <- 19; pch_b <- 1  }

    plot(NULL, xlim = c(a - 0.5, b + 0.5), ylim = c(-0.5, 0.5),
        xlab = "", ylab = "", axes = FALSE)

    # interval line
    segments(a, 0, b, 0, lwd = 2)

    # open circles at endpoints
    points(a, 0, pch = pch_a, cex = 2)
    points(b, 0, pch = pch_b, cex = 2)

    # filled point at x0
    points(x0, 0, pch = 19, cex = 2, col = "red")

    # labels
    text(a,  -0.2, label_a,  cex=1.2)
    text(b,  -0.2, label_b,  cex=1.2)
    text(x0, -0.2, label_x0, cex=1.2, col = "red")
}

epsilon_delta_plot <- function(f, x0, L, epsilon, a, b, f_label, title, zoom=2) {
    delta_left <- abs(x0 - a)
    delta_right <- abs(b - x0)
    delta <- min(delta_left, delta_right)
    xlim <- c(a-delta_left*zoom, b+delta_right*zoom)
    ylim <- c(L-epsilon-epsilon*zoom,L+epsilon+epsilon*zoom)

    x_range <- c(x0 - 3 * delta, x0 + 3 * delta)
    x <- seq(x_range[1], x_range[2], length.out = 500)
    y <- f(x)
    df <- data.frame(x = x, y = y)

    ggplot(df, aes(x, y)) +
      # epsilon band
      annotate("rect",
        xmin = -Inf, xmax = Inf,
        ymin = L - epsilon, ymax = L + epsilon,
        fill = "blue", alpha = 0.1
      ) +
      # delta band
      annotate("rect",
        xmin = x0 - delta, xmax = x0 + delta,
        ymin = -Inf, ymax = Inf,
        fill = "red", alpha = 0.1
      ) +
      geom_line(linewidth = 1) +
      # epsilon lines
      geom_hline(yintercept = L + epsilon, linetype = "dashed", color = "blue") +
      geom_hline(yintercept = L - epsilon, linetype = "dashed", color = "blue") +
      geom_hline(yintercept = L, linetype = "dotted", color = "blue") +
      # delta lines
      geom_vline(xintercept = x0 + delta, linetype = "dashed", color = "red") +
      geom_vline(xintercept = x0 - delta, linetype = "dashed", color = "red") +
      geom_vline(xintercept = x0, linetype = "dotted", color = "red") +
      # labels
      annotate("text", x = x_range[1], y = L + epsilon,
               label = paste("L + e =", L + epsilon), vjust = -1, hjust = 0, color = "blue") +
      annotate("text", x = x_range[1], y = L - epsilon,
               label = paste("L - e =", L - epsilon), vjust =  1, hjust = 0, color = "blue") +
      annotate("text", x = x0 + delta, y = min(y),
               label = paste("x0 + d =", x0 + delta), hjust = 1, color = "red") +
      annotate("text", x = x0 - delta, y = min(y),
               label = paste("x0 - d =", x0 - delta), hjust = 1, color = "red") +
      labs(title = title,
           subtitle = paste("e =", epsilon, "  d =", round(delta, 2)),
           x = "x", y = f_label) +
      theme_minimal() +
      coord_cartesian(xlim = xlim, ylim = ylim)
}

```

# Centering Intervals About a Point

In Exercises 1-6, sketch the interval (a,b) on the x-axis with the point $x_0$
inside. Then find a value of $\delta > 0$ such that for all $x$, $0 < |x - x_0|
< \delta \Rightarrow a < x < b$.

1. a = 1, b = 7, $x_0 = 5$

> $\delta = \min {(|1 - 5|, |7 - 5|)} = \min {(4, 2)} = 2$.

```{r problem 1, out.width="35%", echo=F}
plot_interval(a=1, b=7, x0=5)
```

---

2. a = 1, b = 7, $x_0 = 2$

> $\delta = \min {(|1 - 2|, |7 - 2|)} = \min {(1, 5)} = 5$.

```{r problem 2, out.width="35%", echo=F}
plot_interval(a=1, b=7, x0=2)
```

---

3. a = -7/2, b = -1/2, $x_0 = -3$

> $\delta = \min {(|-7/2 - -3|, |-1/2 - -3|)} = \min {(1/2, 5/2)} = 1/2$.

```{r problem 3, out.width="35%", echo=F}
plot_interval(a=-7/2, b=-1/2, x0=-3)
```

---

4. a = -7/2, b = -1/2, $x_0 = -3/2$

> $\delta = \min {(|-7/2 - -3/2|, |-1/2 - -3/2|)} = \min {(4/2, 1)} = 1$

```{r problem 4, out.width="35%", echo=F}
plot_interval(a=-7/2, b=-1/2, x0=-3/2)
```

---

5. a = 4/9, b = 4/7, $x_0 = 1/2$

> $\delta = \min {(|4/9 - 1/2|, |4/7 - 1/2|)} = \min {(1/18, 1/14)} = 1/18$.

```{r problem 5, out.width="8%", echo=F}
plot_interval(a=4/9, b=4/7, x0=1/2)
```

---

6. a = 2.7591, b = 3.2391, $x_0 = 3$

> $\delta = \min {(|2.7591 - 3|, |3.2391 - 3|)} = \min {(0.2409, 0.2391)} =
  .2391$.

```{r problem 6, out.width="15%", echo=F}
plot_interval(a=2.7591, b=3.2391, x0=3)
```

# Finding Deltas Graphically

In Exercises 7 - 14, use the graphs to find a $\delta > 0$ such that for all $x$
$$ 0 < |x - x_0| < \delta \Rightarrow |f(x) - L| < \epsilon $$

7.

```{r problem 7, echo=F}
epsilon_delta_plot(function(x) 2 * x - 4, x0 = 5, L = 6,
                   epsilon = .2, a = 4.9, b = 5.1, f_label = "f(x)", title = "")
```

$$
\begin{aligned}
f(x)     &= 2x - 4 \\
x_0      &= 5 \\
L        &= 6 \\
\epsilon &= 0.2
\end{aligned}
$$

8.

```{r problem 8, echo=F}
epsilon_delta_plot(
    function(x) -(3/2) * x + 3, x0 = -3, L = 7.5,
    epsilon = .15, a = -3.1, b = -2.9, f_label="f(x)", title="")
```

$$
\begin{aligned}
f(x) &= -\frac{3}{2}x + 3 \\
x_0  &= -3 \\
L    &= 7.5 \\
\epsilon &= 0.15
\end{aligned}
$$

9.

```{r problem 9, echo=F}
epsilon_delta_plot(
    function(x) sqrt(x), x0 = 1, L = 1, epsilon = 1/4, a = 9/16, b = 25/16,
    f_label="f(x)", title="")
```

$$
\begin{aligned}
f(x) &= \sqrt{x} \\
x_0  &= 1 \\
L    &= 1 \\
\epsilon &= \frac{1}{4}
\end{aligned}
$$

10.

```{r problem 10, echo=F}
epsilon_delta_plot(
    function(x) 2 * sqrt(x + 1), x0 = 3, L = 4, epsilon = .2,
    a = 2.61, b = 3.41, f_label="f(x)", title="")
```

$$
\begin{aligned}
f(x) &= 2 \sqrt{x + 1} \\
x_0  &= 3 \\
L    &= 4 \\
\epsilon &= 0.2
\end{aligned}
$$

# Finding Deltas Algebraically

Each of Exercises 15-30 gives a function $f(x)$ and numbers $L$, $x_0$, and
$epsilon > 0$. In each case, find an open interval about $x_0$ on which the
inequality $|f(x) - L| < \epsilon$ holds. Then give a value for $\delta > 0$
such that for all $x$ satisfying $0 < |x - x_0| < \delta$ the inequality $|f(x)
- L| < \epsilon$ holds.

15. $f(x) = x + 1, L = 5, x_0 = 4, \epsilon = 0.01$

$$
\begin{aligned}
|f(x) - L| &< \epsilon \\
|x + 1 - 5| &< 0.01 \\
-0.01 < x - 4 &< 0.01 \\
`r 4 - 0.01` < x &< `r 4 + 0.01`
\end{aligned}
$$

> The inequality holds for the open interval $(`r 4 - 0.01`, `r 4 + 0.01`)$.

> For all $x$ satisfying $0 < |x - x_0| < \delta$, we choose $\delta = \epsilon
  = 0.01$.

---

16. $f(x) = 2x - 2, L = -6, x_0 = -2, \epsilon = 0.02$

$$
\begin{aligned}
|2x - 2| &< 0.02 \\
2|x - 1| &< 0.02 \\
0.01 < x - 1 &< 0.01 \\
`r 1 - 0.01` < x &< `r 1 + 0.01`
\end{aligned}
$$

> The inequality holds for the open interval $(`r 1 - 0.01`, `r 1 + 0.01`)$.

> For all $x$ satisfying $0 < |x - x_0| < \delta$, we choose $\delta =
  \frac{\epsilon}{2} = 0.01$.

---

17. $f(x) = \sqrt{x + 1}, L = 1, x_0 = 0, \epsilon = 0.1$

$$
\begin{aligned}
|\sqrt{x + 1} - 1| < 0.1 \\
-0.1 < \sqrt{x + 1} - 1 < 0.1 \\
`r 1 - 0.1` < \sqrt{x + 1} < `r 1 + 0.1` \\
`r round((1 - .1)^2, 2)` < x + 1 < `r round((1 + .1)^2, 2)` \\
`r round((1 - .1)^2 - 1, 2)` < x < `r round((1 + .1)^2 - 1, 2)`
\end{aligned}
$$

> The inequality holds for the open interval
  $(`r abs(round((1 - .1)^2 - 1, 2))`, `r abs(round((1 + .1)^2 - 1, 2))`)$.

> For all $x$ satisfying $0 < |x - 0| = |x| < \delta$, we choose $\delta = \min
  {(|0 - `r abs(round((1 - .1)^2 - 1, 2))`|, |`r abs(round((1 + .1)^2 - 1, 2))` - 0|)} =
  `r min(abs(0 - round((1 - .1)^2 - 1, 2)), abs(round((1 + .1)^2, 2) - 0))`$.

---

18. $f(x) = \sqrt{x}, L = 1/2, x_0 = 1/4, \epsilon = 0.1$

$$
\begin{aligned}
|\sqrt{x} - 1/2| < 0.1 \\
-0.1 < \sqrt{x} - 1/2< 0.1 \\
.4 < \sqrt{x} < .6 \\
.16 < x < .36
\end{aligned}
$$

> The inequality holds for the open interval $(.16, .36)$.

The distance from $x_0 = 1/4 = .25$ to each endpoint:
$(.25 - .16, .36 - .25) = (.09, .11)$. $\delta = \min {(.09, .11)} = .09$.

# More on Formal Limits

Prove the limit statements in Exercises 37-50.

<!-- Exercise 37 -->

37. $\lim_{x \rightarrow 4} (9 - x) = 5$

> We need to show that for every $\epsilon > 0$, there exists $\delta > 0$ such
that: $$0 < |x - 4| < \delta \Rightarrow |(9 - x) - 5| < \epsilon.$$

> #### Scratch work
  First simplify the expression: $$|(9 - x) - 5| = |4 - x| = |x - 4|.$$

> So we need $|x - 4| < \epsilon$, which means we can simply choose $\delta =
\epsilon$.\newline

> \begin{proof}
Given $\epsilon > 0$, choose $\delta = \epsilon$. Then for all $x$ satisfying $0
< |x - 4| < \delta$ $$|(9 - x) - 5| = |4 - x| = |x - 4| < \delta = \epsilon.$$
Therefore $$\lim_{x \rightarrow 4} (9 - x) = 5.$$
\end{proof}

```{r exercise 37, echo=F}
epsilon <- 0.3
epsilon_delta_plot(
    f = function(x) 9 - x,
    x0 = 4, L = 5,
    epsilon = epsilon,
    a = 4 - epsilon, b = 4 + epsilon,
    f_label = "9 - x",
    title = TeX("$\\lim_{x \\rightarrow 4} (9 - x) = 5$")
)
```

<!-- Exercise 38 -->

38. $\lim_{x \rightarrow 3} (3x - 7) = 2$.

> We need to show that for every $\epsilon > 0$, there exists a $\delta > 0$
  such that: $$0 < |x - 3| < \delta \Rightarrow |(3x - 7) - 2| < \epsilon.$$

> #### Scratch work
$$|(3x - 7) - 2| = |3x - 9| = 3|x - 3|.$$

> Choose $\delta = \epsilon/3$.\newline

> \begin{proof}
Given $\epsilon > 0$, choose $\delta = \epsilon/3$. Then for all $x$ satisfying
$0 < |x - 3| < \delta$ $$|(3x - 7) - 2| = |3x - 9| = 3|x - 3| < 3\delta =
3(\epsilon/3) = \epsilon.$$ Therefore $$\lim_{x \rightarrow 3} (3x - 7) = 2$$.
\end{proof}

```{r exercise 38, echo=F}
epsilon <- 0.3
epsilon_delta_plot(
    f = function(x) 3 * x - 7,
    x0 = 3, L = 2,
    epsilon = epsilon, a = 3 - epsilon / 3, b = 3 + epsilon / 3,
    f_label = "3x - 7",
    title = TeX("$\\lim_{x \\rightarrow 3} 3x - 7 = 2$")
)
```

<!-- Exercise 39 -->

39. $\lim_{x \rightarrow 9} \sqrt{x - 5} = 2$

> We need to show that for every $\epsilon > 0$, there exists a $\delta > 0$
  such that: $$0 < |x - 9| < \delta \Rightarrow |\sqrt{x - 5} - 2| < \epsilon.$$

> #### Scratch work
Simplify the right side by rationalizing:

> $$
\begin{aligned}
|\sqrt{x - 5} - 2| =
  |\sqrt{x - 5} - 2| \cdot \frac{|\sqrt{x - 5} + 2|}{|\sqrt{x - 5} + 2|} =
  \frac{|x - 9|}{|\sqrt{x - 5} + 2}
\end{aligned}
$$

> Since $|\sqrt{x - 5} + 2| \ge 2$ for all $x \ge 5$, we have:

$$
\frac{|x - 9|}{|\sqrt{x - 5} + 2} \le \frac{|x - 9|}{2}
$$

> So we need $\frac{|x - 9|}{2} < \epsilon$, i.e. $|x - 9| < 2\epsilon$. Choose
  $\delta = 2\epsilon$.\newline

> \begin{proof}
Given $\epsilon > 0$, choose $\delta = 2\epsilon$. Then for all $x$ satisfying
$0 < |x - 9| < \delta$ $$|\sqrt{x - 5} - 2| = \frac{|x - 9|}{|\sqrt{x - 5} + 2|}
\le \frac{|x - 9|}{2} < \frac{\delta}{2} = \frac{2\epsilon}{2} = \epsilon.$$
Therefore $$\lim_{x \rightarrow 9} \sqrt{x - 5} = 2.$$
\end{proof}

```{r exercise 39, echo=F}
epsilon <- 0.3
epsilon_delta_plot(
    f = function(x) sqrt(x - 5),
    x0 = 9, L = 2,
    epsilon = epsilon, a = 9 - 2 * epsilon, b = 9 + 2 * epsilon,
    f_label = TeX("$\\sqrt{x - 5}$"),
    title = TeX("$\\lim_{x \\rightarrow 9} \\sqrt{x - 5} = 2$")
)
```

<!-- Exercise 40 -->

40. $\lim_{x \rightarrow 0} \sqrt{4 - x} = 2$

> We need to show that for every $\epsilon > 0$, there exists a $\delta > 0$
  such that: $$0 < |x - 0| = |x| < \delta \Rightarrow |\sqrt{4 - x} - 2| <
  \epsilon.$$

> #### Scratch work
  Simplyfying the right side by rationalizing:

> $$
\begin{aligned}
  |\sqrt{4 - x} - 2| \cdot \frac{|\sqrt{4 - x} + 2|}{|\sqrt{4 - x} + 2|} =
  \frac{|x|}{|\sqrt{4 - x} + 2|}
\end{aligned}
$$

> Since $|\sqrt{4 - x} + 2| \ge 2$ for all $x \le 4$, we have:

$$
\frac{|x|}{|\sqrt{4 - x} + 2|} \le \frac{|x|}{2}
$$

> So we need $\frac{|x|}{2} < \epsilon$, i.e. $|x| < 2\epsilon$. Choose $\delta
  = 2\epsilon$.\newline

> \begin{proof}
Given $\epsilon > 0$, choose $\delta = 2\epsilon$. Then for all $x$ satisfying
$0 < |x| < \delta$ $$ |\sqrt{4 - x} - 2| = \frac{|x|}{|\sqrt{4 - x} + 2|} \le
\frac{|x|}{2} < \frac{\delta}{2} = \frac{2\epsilon}{2} = \epsilon.$$ Therefore
$$\lim_{x \rightarrow 0} \sqrt{4 - x} = 2.$$
\end{proof}

```{r exercise 40, echo=F}
epsilon <- 0.3
epsilon_delta_plot(
    f = function(x) sqrt(4 - x),
    x0 = 0, L = 2,
    epsilon = epsilon, a = 0 - 2 * epsilon, b = 0 + 2 * epsilon,
    f_label = TeX("$\\sqrt{4 - x}$"),
    title = TeX("$\\lim_{x \\rightarrow 0} \\sqrt{4 - x} = 2$")
)
```

<!-- Exercise 41 -->

41. $\lim_{x \rightarrow 1} f(x) = 1$ if $f(x) = \begin{cases} x^2, & x \ne 1 \\
    2, & x = 1\end{cases}$

> We need to show that given $\epsilon > 0$ there exists a $\delta > 0$ such
  that for all $x$ $$0 < |x - 1| < \delta \Rightarrow |f(x) - 1| < \epsilon.$$

> #### Scratch work
Solve the inequality

> $$
\begin{aligned}
|x^2 - 1| < \epsilon &\\
- \epsilon < x^2 - 1 < \epsilon &\\
1 - \epsilon < x^2 < 1 + \epsilon &\\
\sqrt{1 - \epsilon} < |x| < \sqrt{1 + \epsilon}
  & \quad \quad \text{assume }\epsilon < 1 \\
\sqrt{1 - \epsilon} < x < \sqrt{1 + \epsilon}.
\end{aligned}
$$

> The inequality $|f(x) - 1| < \epsilon$ holds for all $x \ne 1$ in the open
interval $(\sqrt{1 - \epsilon}, \sqrt{1 + \epsilon})$.

> We choose $\delta = \min {1 - \sqrt{1 - \epsilon}, \sqrt{1 + \epsilon} -
  1}$.\newline

> \begin{proof}
Given $\epsilon > 0$, assume WLOG $\epsilon < 1$ (if $\epsilon \le 1$ any
$\delta \le 1$ works). Choose: $$\delta = \min{1 - \sqrt{1 - \epsilon}, \sqrt{1
+ \epsilon} - 1}.$$ Then for all $x$ satisfying $0 < |x - 1| < \delta$, we have
 $x \ne 1$ so $f(x) = x^2$. Then:
$$
\begin{aligned}
|x - 1| < \delta \le - \sqrt{1 - \epsilon} \Rightarrow x > \sqrt{1 - \epsilon}
\Rightarrow x^2 > 1 - \epsilon \\\\
|x - 1| < \delta \le \sqrt{1 + \epsilon} + 1 \Rightarrow x < \sqrt{1 + \epsilon}
\Rightarrow x^2 < 1 + \epsilon
\end{aligned}
$$
Combining: $$1 - \epsilon < x^2 < 1 + \epsilon \Rightarrow |x^2 - 1| <
\epsilon$$
Therefore: $$|f(x) - 1| = |x^2 - 1| < \epsilon$$
Hence $\lim_{x \rightarrow 1} f(x) = 1$.
\end{proof}

```{r exercise 41, echo=F}
epsilon <- 0.3
epsilon_delta_plot(
    f = function(x) x^2,
    x0 = 1, L = 1,
    epsilon = epsilon,
    a = 1 - round(1 - sqrt(1 - epsilon), 2), b = 1 + round(1 - sqrt(1 - epsilon), 2),
    f_label = "f(x)",
    title = TeX("$\\lim_{x \\rightarrow 1} f(x) = 1$ if $f(x) = x^2 (x \\ne 1)$ or $2 (x = 1)$")
)
```

<!-- Exercise 42 -->

42. $\lim_{x \rightarrow -2} f(x) = 4 \quad \text{if} \quad f(x) = \begin{cases} x^2, & x \ne -2
    \\ 1, & x = -2\end{cases}$

> We need to show that given $\epsilon > 0$ there exists a $\delta > 0$ such
  that for all $x$ $$0 < |x + 2| < \delta \Rightarrow |f(x) - 4| < \epsilon.$$

> #### Scratch work
Solve the inequality

> $$
\begin{aligned}
|x^2 - 4| < \epsilon &\\
-\epsilon < x^2 - 4 < \epsilon &\\
4 - \epsilon < x^2 < 4 + \epsilon &\\
\sqrt{4 - \epsilon} < x < \sqrt{4 + \epsilon}
    & \quad \text{we want } x \rightarrow -2 \\
- \sqrt{4 + \epsilon} < x < - \sqrt{4 - \epsilon} &\\
\end{aligned}
$$

> The inequality $|f(x) - 4| < \epsilon$ holds for all $x \ne -2$ in the open
  interval $(-\sqrt{4 - \epsilon}, -\sqrt{4 + \epsilon})$.

> We choose

> $$
\begin{aligned}
\delta &= \min {(-2 - (-\sqrt{4 - \epsilon}), -\sqrt{4 - \epsilon} - (-2))} \\
       &= \min {(\sqrt{4 - \epsilon} - 2, 2 - \sqrt{4 - \epsilon})}
\end{aligned}
$$\newline

>\begin{proof}
Given $\epsilon > 0$, assume WLOG $\epsilon < 4$ (if $\epsilon \le$ 4 any $\delta
\le 4$ works). Choose $$\delta = \min {(-2 - \sqrt{4 + \epsilon}, -2 + \sqrt{4 -
\epsilon})}.$$ Then for all $x$ satisfying $0 < |x + 2| < \delta$, we have $x
\ne -2$ so $f(x) = x^2$. Then:
$$
\begin{aligned}
|x + 2| < \delta \le - \sqrt{4 - \epsilon} \Rightarrow x > \sqrt{4 - \epsilon}
\Rightarrow x^2 > 4 - \epsilon\\\\
|x + 2| < \delta \le \sqrt{4 + \epsilon} + 4 \Rightarrow x < \sqrt{4 + \epsilon}
\Rightarrow x^2 < 4 + \epsilon
\end{aligned}
$$
Combining: $$4 - \epsilon < x^2 < 4 + \epsilon \Rightarrow |x^2 - 4| <
\epsilon$$
Therefore: $$|f(x) - 4| = |x^2 - 4| < \epsilon$$
Hence $\lim_{x \rightarrow -2} f(x) = 4$.
\end{proof}

```{r exercise 42, echo=F}
epsilon <- 0.3
epsilon_delta_plot(
    f = function(x) x^2,
    x0 = -2, L = 4,
    epsilon = epsilon,
    a = -2 - round(sqrt(4 + epsilon) - 2, 2), b = -2 + round(sqrt(4 + epsilon) - 2, 2),
    f_label = "f(x)",
    title = TeX("$\\lim_{x \\rightarrow -2} f(x) = 1$ if $f(x) = x^2 (x \\ne 1)$ or $2 (x = 1)$")
)
```

<!-- Exercise 43 -->

43. $\lim_{x \rightarrow 1} \frac{1}{x} = 1$

>  We need to show that given $epsilon > 0$ there exists a $\delta > 0$ such
   that for all $x$ $$0 < |x - 1| < \delta \Rightarrow \left|\frac{1}{x} -
   1\right| < \epsilon$$

> #### Scratch work
Simplify the left side:

$$
\left|\frac{1}{x} - 1 \right| = \left|\frac{1 - x}{x}\right|
  = \frac{|x - 1|}{|x|}
$$

> We need to bound $\frac{1}{|x|}$ from above. Assume $\delta \le 1/2$, so that
  $|x - 1| < 1/2$, which gives:

$$
\frac{1}{2} < x < \frac{3}{2} \Rightarrow |x| > \frac{1}{2} \Rightarrow
  \frac{1}{|x|} < 2
$$

Therefore: $$\frac{|x - 1|}{|x|} < 2|x - 1|$$

So we need $2|x - 1| < \epsilon$, i.e. $|x-1| < \epsilon/2$. Choose:
$$\delta = \min{\left(\frac{1}{2}, \frac{\epsilon}{2}\right)}$$

\begin{proof}
Given $\epsilon > 0$, choose $\delta = \min{\left(\frac{1}{2},
\frac{\epsilon}{2}\right)}.$ Then for all $x$ satisfying $0 < |x - 1| < \delta$:
Since $\delta \le 1/2$ we have $|x - 1| < 1/2$, so $x > 1/2$ and thus
$\frac{1}{|x|} < 2$. Therefore:
$$\left|\frac{1}{x} - 1\right| = \frac{|x - 1|}{|x|} < 2|x-1| < 2\delta
  \le 2 \cdot \frac{\epsilon}{2} = \epsilon$$ Hence $\lim_{x \rightarrow
  1}{\frac{1}{x}} = 1.$
\end{proof}

```{r problem 43, echo=F}
epsilon <- 0.3
epsilon_delta_plot(
    f = function(x) 1/x,
    x0 = 1, L = 1,
    epsilon = epsilon, a = .5, b = epsilon/2,
    f_label = "f(x)",
    title = TeX("$\\lim_{x \\rightarrow 1} \\frac{1}{x} = 1$"),
    zoom = 5
)
```

44. $\lim_{x -> \sqrt{3}}{\frac{1}{x^2}} = 1$

> We need to show that for every $\epsilon > 0$, there exists $\delta > 0$ such
  that:

> $$0 < |x - \sqrt{3}| < \delta \Rightarrow \left|\frac{1}{x^2} -
  \frac{1}{3}\right| < \epsilon$$

> #### Scratch work
Simplify the left side

> $$\left|\frac{1}{x^2} - \frac{1}{3}\right| = \left|\frac{3 - x^2}{3x^2}\right|
    = \frac{|3 - x^2|}{3x^2} = \frac{|x - \sqrt{3}||x + \sqrt{3}|}{3x^2}$$

> We need to bound $\frac{|x + \sqrt{3}|}{3x^2}$ from above. Assume $\delta \le
  1$, so $|x - \sqrt{3}| < 1$, which gives: $$\sqrt{3} - 1 < x < \sqrt{3} + 1$$

> From this we get upper and lower bounds:

> * $|x + \sqrt{3}| < 2 \sqrt{3} + 1$
> * $x > \sqrt{3} - 1 \Rightarrow x^2 > (\sqrt{3} - 1)^2 = 4 - 2 \sqrt{3}
     \Rightarrow \frac{1}{x^2} < \frac{1}{4 - 2\sqrt{3}}$

> Therefore: $$\frac{|x - \sqrt{3}||x + \sqrt{3}|}{3x^2} < 
               \frac{(2\sqrt{3} + 1)}{3(4 - 2\sqrt{3})}|x - \sqrt{3}|$$

> Let $M = \frac{2\sqrt{3} + 1}{3(4 - 2\sqrt{3})}$. So we need $M|x - \sqrt{3}|
  < \epsilon$, i.e. $|x - \sqrt{3}| < \epsilon/M$. Choose:
  $$\delta = \min{\left(1, \frac{\epsilon}{M}\right)}$$

> \begin{proof}
Given $\epsilon > 0$, choose $\delta = \min{\left(1, \frac{\epsilon}{M}\right)}$
where $M = \frac{2\sqrt{3} + 1}{3(4 - 2\sqrt{3})}$. Then for all $x$ satisfying
$0 < |x - \sqrt{3}| < \delta$:  
Since $\delta \le 1$ we have $\sqrt{3} - 1 < x < \sqrt{3} + 1$, so $|x +
\sqrt{3}| < 2\sqrt{3} + 1$ and $x^2 > 4 - 2\sqrt{3}$. Therefore:
$$\left|\frac{1}{x^2}-\frac{1}{3}\right| = \frac{|x - \sqrt{3}||x +
\sqrt{3}|}{3x^2} < \frac{(2\sqrt{3} + 1)}{3(4 - 2\sqrt{3})}|x - \sqrt{3}| <
M\delta \le M \cdot \frac{\epsilon}{M} = \epsilon$$
Hence $\lim_{x \rightarrow \sqrt{3}}{\frac{1}{x^2}} = \frac{1}{3}$.
\end{proof}

---

45. $\lim_{x \rightarrow -3} \frac{x^2 - 9}{x + 3} = -6$

> We need to show that for every $\epsilon > 0$, there exists $\delta > 0$ such
  that:
  $$0 < |x - (-3)| < \delta \Longrightarrow\left|\frac{x^2 - 9}{x + 3} -
     (-6)\right| < \epsilon$$

> #### Scratch work
  Simplify the expression:
  $$\left|\frac{x^2 - 9}{x + 3} + 6\right| = \left|\frac{x^2 + 6x + 9}{x +
  3}\right| = \left|\frac{(x + 3)^2}{x+3}\right| = |x + 3|$$

> We need $|x + 3| < \epsilon$. We can choose $\delta = \epsilon$.\newline

> \begin{proof}
Given $\epsilon > 0$, choose $\delta = \epsilon$. Then for all $x$ satisfying $0
< |x + 3| < \delta$ $$\left|\frac{x^2 - 9}{x+3} + 6\right| = \left|\frac{x^2 +
6x + 9}{x + 3}\right| = |x + 3| < \delta = \epsilon.$$ Therefore $$\lim_{x
\rightarrow -3}\frac{x^2 - 9}{x + 3} = -6.$$
\end{proof}

---

46. $\lim_{x \rightarrow 1} \frac{x^2 - 1}{x - 1} = 2$

> We need to show that for every $\epsilon > 0$, there exists $\delta > 0$ such
  that:
  $$0 < |x - 1| < \delta \Longrightarrow \left|\frac{x^2 - 1}{x - 1} - 2\right|
  < \epsilon$$

> #### Scratch work
  Simplify $$\left|\frac{x^2 - 1}{x - 1} - 2\right| = \left|\frac{(x + 1)(x -
  1)}{x - 1} - 2\right| = |x - 1|$$

> We need $|x - 1| < \epsilon$. We can choose $\delta = \epsilon$.\newline

> \begin{proof}
Give $\epsilon > 0$, choose $\delta = \epsilon$. Then for all $x$ satisfying $0
< |x - 1| < \delta$ $$\left|\frac{x^2 - 1}{x - 1} - 2\right| = \left|\frac{(x -
1)(x + 1)}{x - 1} - 2\right| = |x + 1 - 2| = |x - 1| < \delta = \epsilon.$$
Therefore $$\lim_{x \rightarrow 1} \frac{x^2 - 1}{x - 1} = 2$$.
\end{proof}

# Theory and Examples

51. Define what it means to say that $\lim_{x \rightarrow 2} f(x) = 5$.

> For every $\epsilon > 0$, there exists a $\delta > 0$ such that for all $x$:
  $$0 < |x - 2| < \delta \Longrightarrow |f(x) - 5| < \epsilon$$

> In plain language: we can make $f(x)$ as close to $5$ as we like by taking $x$
  sufficiently close to $2$ (but not equal to $2$).

---

52. Define what it means to say that $\lim_{x \rightarrow 0} g(x) = k$.

> For every $\epsilon > 0$, there exists a $\delta > 0$ such that for all $x$:
  $$0 < |x| < \delta \Longrightarrow |g(x) - k| < \epsilon$$

> In plain language: we can make $g(x)$ as close to $k$ as we like by taking $x$
  sufficiently close to zero (but not equal to zero).

---

53. A wrong statement about limits. Show by example that the following statement
    is wrong.

>> The number $L$ is the limit of $f(x)$ as $x$ approaches $x_0$ if $f(x)$ gets
   closer to $L$ as $x$ approaches $x_0$.

> Explain why the function in your example does not have the given value of $L$
  as a limit as $x \rightarrow x_0$.

>> Take $f(x) = x$ with $x_0 = 0$ and $L = -1$. As $x \rightarrow 0, f(x) = x$
   gets __closer to $-1$__: $$|f(0.1) - (-1)| = 1.1, |f(0.01) - (-1)| = 1.01,
   |f(0.001) - (-1)| = 1.001$$. The values are indeed getting closer to $-1$ as
   $x \rightarrow 0$. But clearly $\lim_{x \rightarrow 0} x = 0 \ne 1$.

>> ### Why the statement is wrong

>> The statement only requires $f(x)$ to get __closer__ to $L$, meaning $|f(x) -
   L|$ decreases as $x \rightarrow x_0$. But "closer" does not mean "close
   enough." The formal definition requires that we can make $|f(x) - L|$
   __arbitrarily small__--less than any $\epsilon > 0$--by taking $x$
   sufficiently close to $x_0$. In the example, $|f(x) - (-1)|$ decreases as $x
   \rightarrow 0$ but never gets below 1, so for $\epsilon = 0.5$ no $\delta$
   can satisfy the formal definition with $L = -1$.
